### 2015-07-02 main.tex
Introduction getting there...

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
% \acmVolume{9}
% \acmNumber{4}
% \acmArticle{39}
% \acmYear{2010}
% \acmMonth{3}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}
%
% % DOI
% \doi{0000001.0000001}
%
% %ISSN
% \issn{1234-56789}

% Document starts
\begin{document}

% Page heads
\markboth{B. Berger et al.}{Title}

% Title portion
\title{Title}
\author{BONNIE BERGER
\affil{Massachusetts Institute of Technology}
\author{NOAH M. DANIELS
\affil{Massachusetts Institute of Technology}
\author{Y. WILLIAM YU
\affil{Massachusetts Institute of Technology}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}
abstract goes here
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}
%
% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

% We no longer use \terms command
%\terms{Design, Algorithms, Performance}

\keywords{computational biology, database search, metagenomics, chemogenomics}

\acmformat{Bonnie Berger, Noah M. Daniels,
and Y. William Yu, 2015. Title.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

\begin{bottomstuff}
% This work is supported by the National Science Foundation, under
% grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
%
% Author's addresses: G. Zhou, Computer Science Department,
% College of William and Mary; Y. Wu  {and} J. A. Stankovic,
% Computer Science Department, University of Virginia; T. Yan,
% Eaton Innovation Center; T. He, Computer Science Department,
% University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
% (Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

\maketitle

\section{Introduction}

Computational biologists answer biological and biomedical 
questions by using computation in support of---or in place of---laboratory 
procedures, in hopes of obtaining more accurate answers at a greatly reduced 
cost.
The past two decades have seen unprecedented technological progress with regard
to generating biological data; next-generation sequencing, mass spectroscopy,
microarrays, nuclear magnetic resonance, and other high-throughput approaches
have led to an explosion of data.
This explosion of data is a mixed blessing.
On the one hand, the scale and scope of data should allow new insights into
genetic and infections diseases, cancer, basic biology, and even human migration
patterns.
On the other hand, researchers are generating datasets so massive that it has 
become difficult to analyze them to discover patterns that give clues to the 
underlying biological processes.

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{moore}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, by some estimates, the amount of genomic data available to 
researchers is increasing by a factor of 10 every year.

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
ALSO MENTION AMAZON AWS HERE.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

cite nature article (Marks)
mention cloud computing, but it doesn't truly address the problem
algorithms for coming up with novel biological insights NOT focus of this review
storage requirements, network transmission, etc.

[image from: http://www.nature.com/nature/journal/v498/n7453/full/498255a.html; picture from EBI]

This Nature article pointed out that currently many omics applications require us to store, access and analyze large libraries of data.
However, the field is realizing that we don’t have the computational tools to tackle these tasks, as Arend Sidow noted.

It’s tempting to think that cloud computing is going to solve this problem.
But that’s not the case.
It might save a little money but doesn’t address the fundamental issue.
It doesn’t change the problem that the data is increasing exponentially faster than computing power per dollar.

Furthermore, as transmission bottlenecks grow as an issue, even getting data into the cloud presents a challenge.

The only solution is to discover fundamentally better algorithms for processing these databases.
Better algorithms can make an enormous difference.

In fact, we need [click] algorithms…
And that’s what we do.
We design algs that do these calculations a lot faster and space-efficiently, and thus their cost doesn’t explode as the size of the databases increase.
* In this talk, we will focus on sublinear-time algorithms, but we also get sublinear-space.

dim reduction - PCA, DCA
indexing - FM-index
kmer approaches (Kingsford)
compressed sensing (Linial)
low-dim polytopes (Alon)
tree decomposition
isorank

Make use of a completely different paradigm for the structure of biological data. By combining redundancy with a low fractal dimension, able to develop algorithms that scale.

Next-gen sequencing, and some applications
Slide 27 to illustrate how sequencing works (2a, 2b)
Slide 28: sequence analysis pipeline (2c)
Redundancy - 
- 1000 genomes project, flybase, wormbase
- subsequence similarity (k-mer approaches)

Exploiting novel structure in biological data
- What data looks like
- actual fractal dimension & metric entropy of example data sets: table
 - metagenomics
 - protein seq
 - bacterial genomes
 - protein structure
 - chemogenomics
- Runtime
- Storage
- Search as exemplar.

Sequence-based data
- cite NRG, see cablast
- cora
- masai, gem
- (include stuff about novelty in mapping over k-mer based approaches from Cora letter to editor)
- existing mappers, cite NRG paper, and pick up where it leaves off
NOT talking about assembly
- metagenomics

Beyond sequence data
- chemogenomics
- protein structure data

Conclusions and future prospect

Other methods from image processing, etc.
Beyond biological data

for other approaches that have NOT been used in biology, see discussion
in conclusion:
other approaches that haven't been used in comp bio: 
- computational geometry (Piotr Indyk)
- succinct data structures
-

sequence approaches

structure approaches

abstraction to distance functions

no coordinate system

these observations about bio data will also hold for data from wildly different sources.

be sure to cite NRG paper, for how to extract structure from high-dimensional data.

% Figure~\ref{fig:one} shows a typical microcontroller C program that
% controls an automotive power window lift. The program is one of the
% programs used in the case study described in Section~\ref{sec:sim}.
% At first sight, the programs looks like an ANSI~C program. It
% contains function calls, assignments, if clauses, and while loops.
% Figure
% \begin{figure}
% \centerline{\includegraphics{acmsmall-mouse}}
% \caption{Code before preprocessing.}
% \label{fig:one}
% \end{figure}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{main}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

\end{document}
% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-07-22 main.tex
Figures included

@@
 
-% Package to generate and customize Algorithm as per ACM style
-\usepackage[ruled]{algorithm2e}
-\renewcommand{\algorithmcfname}{ALGORITHM}
-\SetAlFnt{\small}
-\SetAlCapFnt{\small}
-\SetAlCapNameFnt{\small}
-\SetAlCapHSkip{0pt}
-\IncMargin{-\parindent}
+\usepackage{caption}
+\usepackage{subfig}
 
@@
 \affil{Massachusetts Institute of Technology}
-\author{NOAH M. DANIELS
+NOAH M. DANIELS
 \affil{Massachusetts Institute of Technology}
-\author{Y. WILLIAM YU
-\affil{Massachusetts Institute of Technology}
-% NOTE! Affiliations placed here should be for the institution where the
-%       BULK of the research was done. If the author has gone to a new
-%       institution, before publication, the (above) affiliation should NOT be changed.
-%       The authors 'current' address may be given in the "Author's addresses:" block (below).
-%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
-%       currently affiliated with NASA.
+Y. WILLIAM YU
+\affil{Massachusetts Institute of Technology}}
 
@@
 every year or two; a similar claim can be made about storage capacity 
-(Figure~\ref{moore}).
+(Figure~\ref{fig:growth}).
 In 2002, when the first human genome was sequenced, the growth in computing 
@@
 
+\begin{figure}[htb!]
+\centering
+\subfloat[Growth of genomic sequence data as compared with the combined power of the top-500 supercomputer list; y-axis is logarithmic]{
+\includegraphics[width=3.0in]{assets/moore.png}
+\label{fig:moore}
+}\\
+\subfloat[Growth of genomic sequence data as compared with hard drive capacity; y-axis is logarithmic]{
+\includegraphics[width=3.0in]{assets/kryder.png}
+\label{fig:kryder}
+}
+\caption[Sequencing vs. Storage]{Moore's and Kryder's laws contrasted with genomic sequence data}
+\label{fig:growth}
+\end{figure}
+
 This growth in data poses significant challenges for 
 researchers~\cite{marx2013biology}.
+Currently, many omics applications require us to store, access, and analyze
+large libraries of data.
 One approach to solving these challenges is to embrace cloud computing.
@@
 
-cite nature article (Marks)
-mention cloud computing, but it doesn't truly address the problem
-algorithms for coming up with novel biological insights NOT focus of this review
-storage requirements, network transmission, etc.
+ALGORITHMS FOR COMING UP WITH NOVEL BIOLOGICAL INSIGHTS not FOCUS OF THIS REVIEW
+STORAGE REQUIREMENTS, NETWORK TRANSMISSION, ETC.
 
-[image from: http://www.nature.com/nature/journal/v498/n7453/full/498255a.html; picture from EBI]
+Computer scientists routinely exploit peculiarities in the structure of data in
+order to reduce time or space complexity.
+In computational biology, this approach has served researchers well.
+Now-classical approaches such as principal component analysis (PCA) reduce the 
+dimensionality of data in order to simplify analysis and uncover salient 
+features.
 
-This Nature article pointed out that currently many omics applications require us to store, access and analyze large libraries of data.
-However, the field is realizing that we don’t have the computational tools to tackle these tasks, as Arend Sidow noted.
+\section{Biological sequence data and next-generation sequencing}
 
-It’s tempting to think that cloud computing is going to solve this problem.
-But that’s not the case.
-It might save a little money but doesn’t address the fundamental issue.
-It doesn’t change the problem that the data is increasing exponentially faster than computing power per dollar.
+Many, but not all, of the problems in computational biology deal with sequence
+data, either nucleotide sequences (nominally a four-letter alphabet) or protein
+sequences (nominally a twenty-letter alphabet).
+Certainly, as sequencing is generating the greatest volume of biological data,
+a large portion of bioinformatics algorithms deal primarily with string data.
+For the problem of exact search, approaches for transforming and indexing 
+string data provide benefits.
+The Burrows-Wheeler transform (BWT)~\cite{burrows1994block} provides efficient string 
+compression through a reversible transformation, while the FM-index data 
+structure~\cite{ferragina2000opportunistic} is a compressed substring index, based
+on the BWT, which provides efficient storage as well as fast search.
+Modern next-generation sequencing (NGS) technologies read short fragments of
+DNA, known as \emph{reads}~\ref{fig:shotgun}, which must be mapped onto a reference sequence~\ref{fig:ngs-pipeline}
+BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
+Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for efficient 
+mapping of NGS reads.
 
-Furthermore, as transmission bottlenecks grow as an issue, even getting data into the cloud presents a challenge.
+\begin{figure}[htb!]
+\centering
+\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, or reads, and relies on high coverage to produce a statistically likely representation of a whole genome]{
+\includegraphics[width=3.0in]{assets/shotgun.png}
+\label{fig:shotgun}
+}
+\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
+\includegraphics[width=3.0in]{assets/snps.png}
+\label{fig:snps}
+}\\
+\subfloat[The NGS downstream analysis pipeline.
+Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie.
+The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools.
+This allows relationships between genes and diseases to be uncovered.]{
+\includegraphics[width=3.0in]{assets/ngs-pipeline.png}
+\label{fig:ngs-pipeline}
+}
+\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
+\label{fig:ngs}
+\end{figure}
 
-The only solution is to discover fundamentally better algorithms for processing these databases.
-Better algorithms can make an enormous difference.
+As biological sequence data is string data, it is frequently useful to index or
+count instances of $k$-mers of a fixed length, $k$.
+Search tools such as BLAST~\cite{altschul1990basic} rely on $k$-mer \emph{seeds} in a 
+seed-and-extend approach to sequence search and alignment.
+For counting $k$-mer frequencies, Jellyfish~\cite{marccais2011fast} relies on an efficient
+hash-table encoding as well as Intel's ``compare-and-swap'' CPU instructions for
+improved run-time performance.
 
-In fact, we need [click] algorithms…
-And that’s what we do.
-We design algs that do these calculations a lot faster and space-efficiently, and thus their cost doesn’t explode as the size of the databases increase.
-* In this talk, we will focus on sublinear-time algorithms, but we also get sublinear-space.
+Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
+introduced \emph{compressive genomics}, an approach that relies on compressing
+data in such a way that the desired computation (such as BLAST search) can be
+performed in the compressed representation.
+This approach
 
-dim reduction - PCA, DCA
-indexing - FM-index
-kmer approaches (Kingsford)
-compressed sensing (Linial)
-low-dim polytopes (Alon)
-tree decomposition
-isorank
+(now, redundancy and compressive) loh2012compressive
 
@@
 
+Quality score compression
+
+lossy compression - quartz
+
+Mince~\cite{patro2015data} groups similar reads (those that share a common 
+substring) together into ``buckets'', allowing that common substring to now be 
+removed and treated as the bucket label, so that each read in the compressed 
+representation comprises only its unique differences from the bucket label.
+This allows a general-purpose compressor to achieve tight compression.
+
+\section{Non-sequence data}
+
+Another important type of biological data is gene expression, or the quantity of
+a particular gene product (usually a protein) present in the cell.
+Gene expression data is useful for relating genotype to phenotype, and is 
+important in studying diseases including cancer.
+Usually obtained through microarrays or RNA-seq technologies, expression data is
+quantitative, as each gene from a sample is associated with a numeric expression
+level, and high-dimensional, as many thousands of genes may be analyzed at a
+time.
+While expression data lends itself to cluster analysis and probabilistic 
+approaches, the high dimensionality presents challenges.
+Principal Component Analysis has shown promise in reducing the dimensionality
+of gene expression data~\cite{raychaudhuri2000principal, alter2000singular, yeung2001principal, misra2002interactive}.
+[NEED CITATIONS FOR DIFFICULTIES -- HIGH DIM MICROARRAY DATA]
+Some recent work has explored other ways to exploit the high-dimensional 
+structure of the data.
+SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
+brings ideas from compressed sensing~\cite{candes2005decoding, candes2006stable, donoho2006compressed} 
+to gene expression analysis.
+Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
+are unlikely to be close in a lower-dimensional linear subspace.
+Thus, SPARCLE approximates a gene's expression profile as a linear combination
+of a few other genes' profiles, and then applies supervised learning to
+predict relationships between genes.
+
+Another recent and novel approach to exploiting the structure of gene expression
+space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
+data as a polytope, and infers the specific tasks represented by vertices of
+that polytope from the features most highly enriched at those vertices.
+In human breast tumors, and in mouse tissues, the expression data were
+well-described by tetrahedra whose vertices were enriched for different tumor
+types and biological functions, from which~\cite{hart2015inferring} infer four distinct
+tasks.
+
+Networks are frequently used to represent biological data, such as the genetic
+and physical interactions among proteins, such as those in metabolic pathways.
+Several approaches have taken advantage of the particular topologies of these
+networks, such as diffusion-based approaches, which explore the topology of
+networks through random walks.
+IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn} perform global multiple network 
+alignment based on network topology and conserved biological function.
+IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
+IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
+PageRank-Nibble algorithm~\cite{andersen2006local}.
+In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
+interact with
+many partners; these interactions do not imply shared function as strongly as
+non-hub interactions.
+Diffusion state distance (DSD)~\cite{cao2013going} is another diffusion-based approach
+which defines a distance metric based on random walks.
+DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
+functions when applied to standard approaches such as neighbor voting for function annotation.
+
+As new functional annotations and protein-protein interactions are discovered,
+the number and size of these networks grows.
+Diffusion component analysis (DCA)~\cite{cho} amounts to a network-based 
+analogue of principle component analysis; DCA reduces the dimensionality of a
+network in such a way that function annotation can be performed in the reduced 
+space, significantly reducing running time while still providing competitive
+prediction accuracy.
+
+Finally, in the realm of protein structure prediction, it has been 
+observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
+range of
+possible dihedral angles, protein structures inhabit a smaller subspace, defined
+by a narrower range of angles, referred to as 
+rotamers~\cite{dunbrack1993backbone}.
+The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
+for the
+purpose of sidechain packing (finding the lowest-energy conformation for the
+sidechain of each amino acid), using tree width to reduce the search space and
+simplify dependencies.
+
 Exploiting novel structure in biological data
 - What data looks like
-- actual fractal dimension & metric entropy of example data sets: table
+- actual fractal dimension and metric entropy of example data sets: table
  - metagenomics
@@
 
-% Figure~\ref{fig:one} shows a typical microcontroller C program that
-% controls an automotive power window lift. The program is one of the
-% programs used in the case study described in Section~\ref{sec:sim}.
-% At first sight, the programs looks like an ANSI~C program. It
-% contains function calls, assignments, if clauses, and while loops.
-% Figure
-% \begin{figure}
-% \centerline{\includegraphics{acmsmall-mouse}}
-% \caption{Code before preprocessing.}
-% \label{fig:one}
-% \end{figure}
-
 % Bibliography

### 2015-07-23 main.tex
lousy first draft

@@
 Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
-ALSO MENTION AMAZON AWS HERE.
+Amazon Web Services are also commonly used for computational biology 
+research~\cite{schatz2010cloud}.
 However, while cloud computing frees researchers from maintaining their own
@@
 
-ALGORITHMS FOR COMING UP WITH NOVEL BIOLOGICAL INSIGHTS not FOCUS OF THIS REVIEW
-STORAGE REQUIREMENTS, NETWORK TRANSMISSION, ETC.
-
 Computer scientists routinely exploit peculiarities in the structure of data in
@@
 features.
+This review focuses on algorithmic advances for dealing with the growth in biological data; algorithms for novel biological insights are not its focus.
 
@@
 performed in the compressed representation.
-This approach
-
-(now, redundancy and compressive) loh2012compressive
-
-Make use of a completely different paradigm for the structure of biological data. By combining redundancy with a low fractal dimension, able to develop algorithms that scale.
-
-Next-gen sequencing, and some applications
-Slide 27 to illustrate how sequencing works (2a, 2b)
-Slide 28: sequence analysis pipeline (2c)
-Redundancy - 
-- 1000 genomes project, flybase, wormbase
-- subsequence similarity (k-mer approaches)
-
-Quality score compression
-
-lossy compression - quartz
-
-Mince~\cite{patro2015data} groups similar reads (those that share a common 
+This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
+Coarse search is performed only on the coarse, or representative, subsequences
+that represent unique data.
+Any coarse sequence that is within some threshold of the query is then expanded 
+into all the similar sequences it represents; the fine search is then over this
+(typically small) subset of the original database.
+This approach provides run-time improvements to BLAST search on 
+nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
+data; moreover, these run-time improvements increase as databases grow.
+Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
+similar reads (those that share a common 
 substring) together into ``buckets'', allowing that common substring to now be 
@@
 This allows a general-purpose compressor to achieve tight compression.
+SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
+reads in such a way that a general-purpose compressor achieves improved
+compression ratios.
+
+Along with a genome sequence, modern sequencing systems provide a measure of
+the quality of, or confidence in, the nucleotide base read at every position.
+While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
+bits per position, these quality scores occupy a larger range, na\"ively
+requiring six or seven bits per position.
+Thus, quality scores offer low-hanging fruit with respect to compression for 
+storage and transmission of sequence data.
+Several approaches to compressing quality scores have been 
+presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
+Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
+compression, by bounding the likelihood that a quality score is informative.
+Surprisingly, Quartz not only demonstrates improved compression over competing
+approaches, but slightly improves the accuracy of downstream variant-calling.
+
+The idea of taking advantage of redundancy in sequence data is particularly
+important in the face of new sources of data.
+Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
+variation; instead of producing a single reference genome for a species, many
+complete genomes are catalogued.
+Likewise, WormBase~\cite{harris2014wormbase} and 
+FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
+strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
+respectively.
+On a potentially even larger scale is the growth of metagenomic data.
+Metagenomics is the study of the many genomes (bacterial, fungal, and even 
+viral) that make up a particular environment.
+Such an environment could be soil from a particular region (which can lead to 
+the discovery of new 
+antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
+the human gut, whose microbiome has been linked to human-health concerns 
+including Autism Spectrum Disorder~\cite{macfabe2012short}, 
+Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
+obesity~\cite{greenblum2012metagenomic}.
+
+The continued ability to search and analyze these growing data sets hinges on
+clever algorithms that take advantage of the structure of, and redundancy 
+present in, the data.
+Indeed, these growing data sets ``threaten to make the arising problems 
+computationally infeasible.''~\cite{berger2013computational}
+
+One approach to more quickly searching these data sets is to come up with an
+indexing technique that leverages domain knowledge about the structure of the
+data.
+In the realm of NGS read mapping, the Genome Multitool (GEM)
+mapper~\cite{marco2012gem} makes use of a Ferragina-Manzini 
+index~\cite{ferragina2000opportunistic} coupled with dynamic programming in
+a compressed representation, in order to prune the search space when mapping
+reads to a reference genome.
+Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
+the space of possible matches, likewise pruning the search space.
+
+Another approach is to make use of a completely different paradigm for the 
+structure of biological data. 
+By combining redundancy with a low fractal 
+dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
+scale even as the data continue to grow.
+The CORA read mapper~\cite{yorukoglu2015compressive} applies a $k$-mer based 
+read-compression approach with a compressive indexing of the reference genome
+(referred to as a homology table).
+Short seed-clustering schemes, such as those used in Masai and 
+MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
+schemes aim to accelerate only the seed-to-reference matching step.
+Thus, there is a subsequent seed-extension step, which is substantially more 
+costly and still needs to be performed for each read and mapping individually, 
+even when seeds are clustered. 
+Through its k-mer based read compression model, CORA is able to accelerate and 
+achieve asymptotically sublinear scaling for both the seed-matching and 
+seed-extension steps within coarse-mapping, which comprises the major bulk of 
+the read-mapping computation.
+
+In the realm of metagenomic search, BLASTX~\cite{altschul1990basic} is commonly
+used to translate nucleotide reads into their possible protein sequences, and
+search for them in a protein database.
+Recent advances in metagenomic search tools have relied on three advances over
+BLASTX: indexing, alphabet reduction, and compression.
+RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
+collision-free hash table.
+The alphabet reduction, as it is reversible, can be thought of as a form of
+lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
+alphabet, with offsets stored to recover the original sequence in the full
+alphabet.
+The hash-table provides an efficient index of the database to be searched.
+DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
+``shaped seeds'' instead of simple $k$-mer seeds to index the database.
+
+The recently-released MICA~\cite{yu2015entropy} demonstrates that the 
+compressive-acceleration approach of 
+caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
+largely orthogonal to alphabet-reduction and indexing approaches.
+MICA applies the compressive-acceleration framework to DIAMOND, using DIAMOND
+for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
+``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
+caBLASTP does over BLASTP.
 
@@
 While expression data lends itself to cluster analysis and probabilistic 
-approaches, the high dimensionality presents challenges.
+approaches, the high dimensionality presents challenges~\cite{nguyen2002tumor}.
 Principal Component Analysis has shown promise in reducing the dimensionality
-of gene expression data~\cite{raychaudhuri2000principal, alter2000singular, yeung2001principal, misra2002interactive}.
-[NEED CITATIONS FOR DIFFICULTIES -- HIGH DIM MICROARRAY DATA]
+of gene expression data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
 Some recent work has explored other ways to exploit the high-dimensional 
@@
 SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
-brings ideas from compressed sensing~\cite{candes2005decoding, candes2006stable, donoho2006compressed} 
+brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
 to gene expression analysis.
@@
 the number and size of these networks grows.
-Diffusion component analysis (DCA)~\cite{cho} amounts to a network-based 
+Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
 analogue of principle component analysis; DCA reduces the dimensionality of a
@@
 
-Finally, in the realm of protein structure prediction, it has been 
+In the realm of protein structure prediction, it has been 
 observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
@@
 
-Exploiting novel structure in biological data
-- What data looks like
-- actual fractal dimension and metric entropy of example data sets: table
- - metagenomics
- - protein seq
- - bacterial genomes
- - protein structure
- - chemogenomics
-- Runtime
-- Storage
-- Search as exemplar.
+The compressive acceleration approach of \cite{loh2012compressive} has been
+generalized and adapted to non-sequence spaces as well.
+Chemogenomics, the computational study of small-molecule compounds and their
+affect on phenotype, relies on comparing chemical graph structures to identify
+similar molecules and binding sites.
+However, there are once again an increasing number of such chemical compounds
+to search; the NCBI's PubChem database has grown from 31 million compounds in
+January 2011 to 68 million in July, 2015.
+Furthermore, comparing chemical graph structures typically involves computing
+the maximal common subgraph (MCS), an NP-hard problem.
+One approach to this type of search is the Small Molecule Subgraph Detector 
+(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
+on the size and complexity of the graphs in question.
+Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
+accelerates SMSD search by an average of 150x on the PubChem database.
 
-Sequence-based data
-- cite NRG, see cablast
-- cora
-- masai, gem
-- (include stuff about novelty in mapping over k-mer based approaches from Cora letter to editor)
-- existing mappers, cite NRG paper, and pick up where it leaves off
-NOT talking about assembly
-- metagenomics
+When the structure of a protein is known but its biological function or 
+evolutionary relationships are not, researchers may search for structurally
+similar proteins that are better studied~\cite{gibrat1996surprising}.
+Classical tools for this involve performing pairwise structural alignments to
+look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
+along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
+Matt~\cite{menke2008matt}.
+FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
+approximating structural alignments, which can be computationally demanding,
+by instead comparing the ``bag-of-words'' from each structure.
+Analogous to a term-frequency vector in information retrieval, this bag-of-words
+indicates the abundance of particular, short structural motifs within a protein.
+Again applying a compressive acceleration approach, 
+esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
+Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
+running time by an average of 10x.
 
-Beyond sequence data
-- chemogenomics
-- protein structure data
+\section{Conclusions and future prospects}
 
-Conclusions and future prospect
+The explosion of biological data, largely due to technological advances such as
+next-generation sequencing, presents us with challenges as well as 
+opportunities.
+The promise of unlocking the secrets of diseases such as cancer, obesity,
+Alzheimer's, autism spectrum disorder, and many others, as well as better 
+understanding the basic science of biology, relies on researchers' ability to 
+analyze the growing flood of genomic, metagenomic, structural, and interactome 
+data.
 
-Other methods from image processing, etc.
-Beyond biological data
+The approach of compressive acceleration~\cite{loh2012compressive}, and its 
+demonstrated ability to scale with the metric entropy of the 
+data~\cite{yu2015entropy}, while providing orthogonal benefits to many other
+useful indexing techniques, is an important tool for coping with the deluge of
+data.
+The extension of this compressive acceleration approach to 
+metagenomics~\cite{yu2015entropy}, NGS read 
+mapping and chemogenomics~\cite{yorukoglu2015compressive} suggests its 
+flexibility.
 
-for other approaches that have NOT been used in biology, see discussion
-in conclusion:
-other approaches that haven't been used in comp bio: 
-- computational geometry (Piotr Indyk)
-- succinct data structures
--
+The field of computational biology must continue to innovate, but also to 
+incorporate the best ideas from other areas of computer science.
+For example, the compressive-acceleration approach bears similarity to a metric 
+ball tree, first described in the database community over twenty years 
+ago~\cite{uhlmann1991satisfying}.
+Other ideas from image processing, computational geometry, and other areas 
+outside of biology are likely to bear fruit.
+It is also likely that algorithmic ideas developed within computational biology
+will become useful in other fields experiencing a data deluge, such as astronomy
+or social networks~\cite{stephens2015big}.
 
-sequence approaches
+While this Review has focused on the importance of algorithms that leverage the
+structure of data to cope with the increased throughput in computational 
+biology, it is also essential that techniques continue to make better 
+predictions and models that demonstrate improved fidelity to wet-lab 
+experiments.
 
-structure approaches
-
-abstraction to distance functions
-
-no coordinate system
-
-these observations about bio data will also hold for data from wildly different sources.
-
-be sure to cite NRG paper, for how to extract structure from high-dimensional data.
+\begin{table}
+\caption{Metric entropy ratio (ratio of clusters to entries in database) and
+fractal dimension at typical search radii for four biological data sets.
+NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
+\begin{tabular}{ccc}
+\hline
+Data set & Metric entropy ratio & Fractal dimension \\        
+\hline
+Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
+\hline
+Protein sequences (NCBI NR) & 5:1 & 1.6\\
+\hline
+Protein structure (PDB) & 10:1 & 2.5\\
+\hline
+Chemical structure (PubChem) & 11:1 & 0.2\\
+\hline
+\end{tabular}
+\end{table}
 
 % Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
+% \bibliographystyle{ACM-Reference-Format-Journals}
+\bibliographystyle{abbrv}
 \bibliography{main}
-                             % Sample .bib file with references that match those in
-                             % the 'Specifications Document (V1.5)' as well containing
-                             % 'legacy' bibs and bibs with 'alternate codings'.
-                             % Gerry Murray - March 2012
 

### 2015-07-26 main.tex
bonnie's edits and reorg

@@
 one thousand U.S. Dollars.
-As a result, by some estimates, the amount of genomic data available to 
+As a result, the amount of genomic data available to 
 researchers is increasing by a factor of 10 every year.
@@
 \centering
-\subfloat[Growth of genomic sequence data as compared with the combined power of the top-500 supercomputer list; y-axis is logarithmic]{
+\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
 \includegraphics[width=3.0in]{assets/moore.png}
@@
 }\\
-\subfloat[Growth of genomic sequence data as compared with hard drive capacity; y-axis is logarithmic]{
+\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
 \includegraphics[width=3.0in]{assets/kryder.png}
@@
 }
-\caption[Sequencing vs. Storage]{Moore's and Kryder's laws contrasted with genomic sequence data}
+\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
 \label{fig:growth}
@@
 dimensionality of data in order to simplify analysis and uncover salient 
-features.
-This review focuses on algorithmic advances for dealing with the growth in biological data; algorithms for novel biological insights are not its focus.
+features~\cite{berger2013computational}.
+As another example, clever indexing techniques such as the Burroughs-Wheeler 
+Transform (BWT) also take advantage of aspects of sequence 
+data~\cite{berger2013computational}.
+This review focuses on algorithmic advances for dealing with the growth in 
+biological data; algorithms for gaining novel biological insights are not its 
+focus.
 
-\section{Biological sequence data and next-generation sequencing}
+\section{Types and sources of biological data}
 
-Many, but not all, of the problems in computational biology deal with sequence
-data, either nucleotide sequences (nominally a four-letter alphabet) or protein
-sequences (nominally a twenty-letter alphabet).
-Certainly, as sequencing is generating the greatest volume of biological data,
-a large portion of bioinformatics algorithms deal primarily with string data.
-For the problem of exact search, approaches for transforming and indexing 
-string data provide benefits.
-The Burrows-Wheeler transform (BWT)~\cite{burrows1994block} provides efficient string 
+In the central dogma of molecular biology, DNA is transcribed into RNA, which
+is translated by the ribosome into a polypeptide chain, a sequence of amino 
+acids, known as a protein.
+This protein folds into a complex, low-energy structure, and functions as
+a cellular machine itself; the DNA sequence determines the amino acid sequence,
+which in turn determines the folded structure of the protein.
+This structure ultimately determines the protein's function within the cell.
+Certain kinds of RNA also function, sometimes with proteins, as cellular 
+machines.
+Thus, researchers are interested in DNA, RNA, and protein sequences, as well
+as the structures that RNA and proteins form.
+Researchers also study the resulting functions of RNA and proteins, in what
+abundance they occur (this is known as gene expression), and how they interact
+with each other (interaction networks).
+
+Most but not all biological data take the form of \emph{sequence} data, either 
+nucleotide sequences (nominally a four-letter alphabet representing the four 
+DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
+representing the twenty standard amino acids).
+Sequence data are obtained in several ways, including mass spectrometry (which
+can determine protein sequence) and RNA-seq (which can determine RNA sequence, 
+and by inference the protein sequence into which it might translate).
+However, the greatest volume of sequence data derive from next-generation 
+sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
+At the dawn of the genomic era, Sanger sequencing was the most widely used 
+method.
+Slow and labor intensive but well-tested, Sanger sequencing is still used to 
+validate NGS results.
+More recently, Sanger sequencing has largely been supplanted by NGS approaches,
+beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
+greater throughput due to massive parallelism, low cost, and simple preparation
+of samples.
+Importantly to this Review, Illumina sequencing, as well as other approaches
+such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
+molecule end-to-end as one could a file from a hard drive.
+Instead, DNA molecules are chopped into many small fragments, or 
+\emph{reads} (Figure ~\ref{fig:shotgun}), 
+which are read concurrently, and which must be reassembled in the correct order 
+if an entire genome is to be read.
+Reads most typically range from 50 to 200 bases long, though longer reads are
+available with some technologies.
+This so-called ``shotgun sequencing'' is analogous to the problem of 
+reconstructing a large file from small, 
+randomly-ordered reads, or a shredded paper document.
+However, these NGS reads overlap, and they are many in number, so 
+reconstructing an entire genome is feasible.
+No sequencing technology is completely infallible, so sequencing machines also
+provide a \emph{quality score} (or measure of the confidence in the DNA base
+called) associated with each position.
+Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
+characters that encode the quality of the base call.
+A sequencing run will produce many of these reads, typically stored in a FASTQ
+file.
+
+Protein structure, primarily determined through X-ray crystallography or
+nuclear magnetic resonance (NMR), describes the coordinates of each atom, in 
+Angstroms, of a protein in three-dimensional space.
+This information is important to understanding how a protein functions, because
+proteins interact physically with other molecules.
+Protein structure data are commonly stored in the Protein Data Bank (PDB); 
+each PDB entry has an associated resolution indicating the accuracy of the
+structural description.
+Thus a PDB entry at its most basic level contains a list of atoms, along with
+which amino acid they belong to and their spatial coordinates as floating-point
+numbers.
+
+Chemical compounds represent another source of structural data, more general
+than proteins.
+Determined by X-ray crystallography, NMR, electron microscopy and other 
+techniques, compounds---many of which are useful pharmaceuticals---can be
+represented as labeled graphs.
+In such a graph, the vertices are atoms (labeled with chemical element) and
+the edges are bonds between atoms (labeled with bond type, such as single or
+double).
+Chemical compounds are stored in databases such as PubChem, and can be described
+as adjacency matrices.
+
+\section{Challenges in biological data}
+
+Given DNA or RNA reads from NGS technologies, the first task is to 
+\emph{assemble} those fragments of sequence (which have associated quality 
+scores) into contiguous sequences.
+The task of \emph{de novo} assembly is beyond the scope of this Review, but is
+possible because reads overlap and because the sequence is covered by many such
+overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
+commonly-used data structure.
+Often, however, a \emph{reference genome} (or in the case of RNA, 
+\emph{transcriptome}) is available for the organism being sequenced; the
+establishment of a human reference genome was indeed the purpose of the Human
+Genome Project.
+When a reference sequence is available, NGS reads must be \emph{mapped} onto
+this reference (Figure ~\ref{fig:ngs-pipeline}).
+This allows the differences between the newly-sequenced genome and the reference
+to be analyzed; these differences may include single-nucleotide polymorphisms
+(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
+insertions 
+or deletions, or larger-scale changes in the genome.
+While reference-based read mapping is a fundamentally simpler problem than
+\emph{de novo} assembly, it is still computationally complex, as gigabytes or
+terabytes of reads must each be mapped onto the reference genome, which can 
+range from millions (for bacteria) to billions (for mammals).
+
+Standard techniques for reference-based read mapping rely on algorithmic 
+approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
+which provides efficient string 
 compression through a reversible transformation, while the FM-index data 
-structure~\cite{ferragina2000opportunistic} is a compressed substring index, based
-on the BWT, which provides efficient storage as well as fast search.
-Modern next-generation sequencing (NGS) technologies read short fragments of
-DNA, known as \emph{reads}~\ref{fig:shotgun}, which must be mapped onto a reference sequence~\ref{fig:ngs-pipeline}
+structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
+based on the BWT, which provides efficient storage as well as fast search.
 BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
-Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for efficient 
-mapping of NGS reads.
+Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
+efficient mapping of NGS reads.
+The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
+coupled with dynamic programming in a compressed representation, in order to 
+prune the search space when mapping reads to a reference genome.
+Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
+the space of possible matches, likewise pruning the search space.
+Even with these approaches, read mapping remains a significant bottleneck in
+genomic research~\cite{hach2010mrsfast}.
 
@@
 
-As biological sequence data is string data, it is frequently useful to index or
-count instances of $k$-mers of a fixed length, $k$.
-Search tools such as BLAST~\cite{altschul1990basic} rely on $k$-mer \emph{seeds} in a 
-seed-and-extend approach to sequence search and alignment.
-For counting $k$-mer frequencies, Jellyfish~\cite{marccais2011fast} relies on an efficient
-hash-table encoding as well as Intel's ``compare-and-swap'' CPU instructions for
-improved run-time performance.
+Given a sequenced genome, the next natural questions are what genes (genomic
+regions that code for proteins) are present, what structure each resulting
+protein takes, and what biological function it performs.
+Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
+beyond the scope of this Review.
+However, determining evolutionary relationships, structure, and function are
+at the heart of current research in computational biology.
+Since some organisms (known as \emph{model organisms}) are better-studied than
+others, and evolution is known to conserve sequence, structure, and function,
+a powerful approach to determine these attributes is to search for similar
+sequences about which more is known.
+This so-called \emph{homology search} entails searching for \emph{approximate}
+matches in databases of known gene or protein sequences.
 
-Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
-introduced \emph{compressive genomics}, an approach that relies on compressing
-data in such a way that the desired computation (such as BLAST search) can be
-performed in the compressed representation.
-This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
-Coarse search is performed only on the coarse, or representative, subsequences
-that represent unique data.
-Any coarse sequence that is within some threshold of the query is then expanded 
-into all the similar sequences it represents; the fine search is then over this
-(typically small) subset of the original database.
-This approach provides run-time improvements to BLAST search on 
-nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
-data; moreover, these run-time improvements increase as databases grow.
-Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
-similar reads (those that share a common 
-substring) together into ``buckets'', allowing that common substring to now be 
-removed and treated as the bucket label, so that each read in the compressed 
-representation comprises only its unique differences from the bucket label.
-This allows a general-purpose compressor to achieve tight compression.
-SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
-reads in such a way that a general-purpose compressor achieves improved
-compression ratios.
+Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
+the standard too for performing homology search on databases of nucleotide and
+protein sequences.
+BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
+that might lead to longer matches, and greedily extends them, ultimately 
+producing a \emph{sequence alignment} between a query and each potential 
+database hit.
+However, BLAST's running time is linear in the size of the database being
+searched, which is problematic as sequence databases continue to grow at a
+faster rate than Moore's law.
 
-Along with a genome sequence, modern sequencing systems provide a measure of
-the quality of, or confidence in, the nucleotide base read at every position.
-While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
-bits per position, these quality scores occupy a larger range, na\"ively
-requiring six or seven bits per position.
-Thus, quality scores offer low-hanging fruit with respect to compression for 
-storage and transmission of sequence data.
-Several approaches to compressing quality scores have been 
-presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
-Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
-compression, by bounding the likelihood that a quality score is informative.
-Surprisingly, Quartz not only demonstrates improved compression over competing
-approaches, but slightly improves the accuracy of downstream variant-calling.
-
-The idea of taking advantage of redundancy in sequence data is particularly
-important in the face of new sources of data.
 Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
@@
 
-The continued ability to search and analyze these growing data sets hinges on
-clever algorithms that take advantage of the structure of, and redundancy 
-present in, the data.
-Indeed, these growing data sets ``threaten to make the arising problems 
-computationally infeasible.''~\cite{berger2013computational}
-
-One approach to more quickly searching these data sets is to come up with an
-indexing technique that leverages domain knowledge about the structure of the
-data.
-In the realm of NGS read mapping, the Genome Multitool (GEM)
-mapper~\cite{marco2012gem} makes use of a Ferragina-Manzini 
-index~\cite{ferragina2000opportunistic} coupled with dynamic programming in
-a compressed representation, in order to prune the search space when mapping
-reads to a reference genome.
-Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
-the space of possible matches, likewise pruning the search space.
-
-Another approach is to make use of a completely different paradigm for the 
-structure of biological data. 
-By combining redundancy with a low fractal 
-dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
-scale even as the data continue to grow.
-The CORA read mapper~\cite{yorukoglu2015compressive} applies a $k$-mer based 
-read-compression approach with a compressive indexing of the reference genome
-(referred to as a homology table).
-Short seed-clustering schemes, such as those used in Masai and 
-MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
-schemes aim to accelerate only the seed-to-reference matching step.
-Thus, there is a subsequent seed-extension step, which is substantially more 
-costly and still needs to be performed for each read and mapping individually, 
-even when seeds are clustered. 
-Through its k-mer based read compression model, CORA is able to accelerate and 
-achieve asymptotically sublinear scaling for both the seed-matching and 
-seed-extension steps within coarse-mapping, which comprises the major bulk of 
-the read-mapping computation.
-
-In the realm of metagenomic search, BLASTX~\cite{altschul1990basic} is commonly
+Metagenomics fundamentally asks what organisms are present, and (in the case
+of a microbiome such as the gut) what metabolic functions it can accomplish as
+a whole.
+One approach to this is to attempt to map NGS reads from a metagenomic sample
+onto a set of reference genomes that are expected to be present.
+This is exactly the read-mapping problem discussed above, but with many 
+reference genomes, compounding the computational requirements.
+A second approach is to perform homology search on a protein sequence database;
+exact or nearly-exact matches imply the presence of a species, while more 
+distant hits may still give clues to function.
+For this task, BLASTX~\cite{altschul1990basic} is commonly
 used to translate nucleotide reads into their possible protein sequences, and
 search for them in a protein database.
-Recent advances in metagenomic search tools have relied on three advances over
-BLASTX: indexing, alphabet reduction, and compression.
+Recent advances in metagenomic search tools have relied on two advances over
+BLASTX: indexing and alphabet reduction.
 RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
@@
 ``shaped seeds'' instead of simple $k$-mer seeds to index the database.
+DIAMOND demonstrates search performance three to four orders of magnitude faster
+than BLASTX, but it is still linear in the size of the database being searched.
 
-The recently-released MICA~\cite{yu2015entropy} demonstrates that the 
-compressive-acceleration approach of 
-caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
-largely orthogonal to alphabet-reduction and indexing approaches.
-MICA applies the compressive-acceleration framework to DIAMOND, using DIAMOND
-for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
-``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
-caBLASTP does over BLASTP.
-
-\section{Non-sequence data}
+Over the course of evolutionary time, structure is known to be more highly 
+conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
+does not change as rapidly.
+When the structure of a protein is known but its biological function or 
+evolutionary relationships are not, researchers may search for structurally
+similar proteins that are better studied~\cite{gibrat1996surprising}.
+Classical tools for this involve performing pairwise structural alignments to
+look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
+along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
+Matt~\cite{menke2008matt}.
+FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
+approximating structural alignments, which can be computationally demanding,
+by instead comparing the ``bag-of-words'' from each structure.
+Analogous to a term-frequency vector in information retrieval, this bag-of-words
+indicates the abundance of particular, short structural motifs within a protein.
 
@@
 
-In the realm of protein structure prediction, it has been 
-observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
-range of
-possible dihedral angles, protein structures inhabit a smaller subspace, defined
-by a narrower range of angles, referred to as 
-rotamers~\cite{dunbrack1993backbone}.
-The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
-for the
-purpose of sidechain packing (finding the lowest-energy conformation for the
-sidechain of each amino acid), using tree width to reduce the search space and
-simplify dependencies.
+The continued ability to search and analyze these growing data sets hinges on
+clever algorithms that take advantage of the structure of, and redundancy 
+present in, the data.
+Indeed, these growing data sets ``threaten to make the arising problems 
+computationally infeasible.''~\cite{berger2013computational}
+
+\section{Compressive omics}
+
+Another approach is to make use of a completely different paradigm for the 
+structure of biological data.
+Much biological data is highly redundant; if a computation is performed on one
+human genome, and a researcher wishes to perform the same computation on another
+human genome, most of the work has already been done.
+
+Seeking to take advantage of the redundancy inherent in genomic sequence data, 
+\cite{loh2012compressive}
+introduced \emph{compressive genomics}, an approach that relies on compressing
+data in such a way that the desired computation (such as BLAST search) can be
+performed in the compressed representation.
+This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
+Coarse search is performed only on the coarse, or representative, subsequences
+that represent unique data.
+Any coarse sequence that is within some threshold of the query is then expanded 
+into all the similar sequences it represents; the fine search is then over this
+(typically small) subset of the original database.
+This approach provides run-time improvements to BLAST search on 
+nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
+data; moreover, these run-time improvements increase as databases grow.
+
+The CORA read mapper~\cite{yorukoglu2015compressive} applies an $l$-mer based 
+read-compression approach with a compressive indexing of the reference genome
+(referred to as a homology table).
+CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
+mappers including BWA, and Bowtie2) by allowing them to operate in a compressed
+space, and relying on a coarse and a fine phase.
+Short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
+MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
+schemes aim to accelerate only the seed-to-reference matching step.
+Thus, there is a subsequent seed-extension step, which is substantially more 
+costly and still needs to be performed for each read and mapping individually, 
+even when seeds are clustered. 
+Through its k-mer based read compression model, CORA is able to accelerate and 
+achieve asymptotically sublinear scaling for both the seed-matching and 
+seed-extension steps within coarse-mapping, which comprises the major bulk of 
+the read-mapping computation.
+
+NEED TO DISCUSS $l$-mers here.
+
+In the area of metagenomic search,
+the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
+compressive-acceleration approach of 
+caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
+largely orthogonal to alphabet-reduction and indexing approaches.
+MICA applies the compressive-acceleration framework to 
+DIAMOND~\cite{buchfink2014fast}, using DIAMOND
+for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
+``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
+caBLASTP does over BLASTP.
 
@@
 
-When the structure of a protein is known but its biological function or 
-evolutionary relationships are not, researchers may search for structurally
-similar proteins that are better studied~\cite{gibrat1996surprising}.
-Classical tools for this involve performing pairwise structural alignments to
-look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
-along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
-Matt~\cite{menke2008matt}.
-FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
-approximating structural alignments, which can be computationally demanding,
-by instead comparing the ``bag-of-words'' from each structure.
-Analogous to a term-frequency vector in information retrieval, this bag-of-words
-indicates the abundance of particular, short structural motifs within a protein.
 Again applying a compressive acceleration approach, 
@@
 running time by an average of 10x.
+
+\section{Structure of biological data}
+
+Cluster-based search is well-studied~\cite{jardine1971use}, but conventional
+wisdom holds that it provides a constant factor speed-up over exhaustive search.
+However, \cite{daniels2013compressive} observed that caBLASTP demonstrated
+increased gains over BLASTP on newer, larger protein databases, suggesting that
+compressive acceleration takes advantage of ever-increasing redundancy in the
+data.
+Beyond redundancy, however, another attribute of large biological data sets
+stands out.
+Any high-dimensional data set, such as amino acid sequences of a bounded 
+length, can be described as inhabiting a vector space.
+For example, given an alphabet of 20 amino acids and a length bounded at 
+500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
+in length, proteins are typically described as comprising \emph{domains} that 
+evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
+could
+enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
+of July 
+2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
+This is not simply due to undersampling; thanks to evolution, only those protein
+sequences which exhibit useful biological function should survive, and most
+random sequences of amino acids would not be expected to form stable structures.
+Thus, not only do biological data exhibit redundancy, they also tend not to 
+inhabit the entire feasible space.
+It seems that physical laws have constrained the data to a particular subspace
+of the enumerable Cartesian space.
+
+One key insight as to why the ``compressive omics'' approach allows computation
+that scales sublinearly in the size of the data is that such data sets exhibit
+low \emph{metric entropy}~\cite{yu2015entropy}.
+That is, for a given cluster radius $r_c$ and a database $D$, the number 
+$N_{r_c} (D)$ of clusters needed to cover $D$ is relatively small compared to
+$|D|$, the number of entries in the database.
+In contrast, if the points were uniformly distributed about the Cartesian space,
+$N_{r_c} (D)$ would be larger.
+
+A second key insight is that the biological data sets considered 
+by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
+That is, for some range of radii $r_1$ and $r_2$,
+
+discrete?
+
+By combining redundancy with a low fractal 
+dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
+scale even as the data continue to grow.
+
+wherever physical or empirical laws have constrained data to 
+a subspace of low entropy and fractal dimension.
+
+Specifically, the ratio $\frac{|D|}{k}$ provides an estimate of the acceleration factor
+for just the coarse search component compared to a full linear search.
+Local fractal dimension around a data point can be computed by determining the
+number of other data points within two radii $r_1$ and $r_2$ of that point;
+given those point counts ($n_1$ and $n_2$, respectively), fractal dimension $d$
+is simply $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$.
+Sampling this property over a dataset can provide a global average fractal 
+dimension.
+When we search a larger radius around a query, the number of points we encounter grows exponentially with the fractal dimension;
+low fractal dimension implies that this growth will not obviate the gains provided by an entropy-scaling data structure.
+
+An entropy-scaling search tree is also related to a metric ball tree~\cite{uhlmann1991satisfying}, although with different time complexity.
+Querying a metric ball tree requires $O(\log n)$ time, assuming the relatively uniform distribution of data points in a metric space.
+This is different from the nonuniform distribution under which entropy-scaling
+search behaves well.
+
+\begin{table}
+\caption{Metric entropy ratio (ratio of clusters to entries in database) and
+fractal dimension at typical search radii for four biological data sets.
+NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
+\begin{tabular}{ccc}
+\hline
+Data set & Metric entropy ratio & Fractal dimension \\        
+\hline
+Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
+\hline
+Protein sequences (NCBI NR) & 5:1 & 1.6\\
+\hline
+Protein structure (PDB) & 10:1 & 2.5\\
+\hline
+Chemical structure (PubChem) & 11:1 & 0.2\\
+\hline
+\end{tabular}
+\end{table}
+
+\section{Storage compression}
+
+Along with a genome sequence, modern sequencing systems provide a measure of
+the quality of, or confidence in, the nucleotide base read at every position.
+While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
+bits per position, these quality scores occupy a larger range, na\"ively
+requiring six or seven bits per position.
+Thus, quality scores offer low-hanging fruit with respect to compression for 
+storage and transmission of sequence data.
+Several approaches to compressing quality scores have been 
+presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
+Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
+compression, by bounding the likelihood that a quality score is informative.
+Surprisingly, Quartz not only demonstrates improved compression over competing
+approaches, but slightly improves the accuracy of downstream variant-calling.
+
+Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
+similar reads (those that share a common 
+substring) together into ``buckets'', allowing that common substring to now be 
+removed and treated as the bucket label, so that each read in the compressed 
+representation comprises only its unique differences from the bucket label.
+This allows a general-purpose compressor to achieve tight compression.
+SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
+reads in such a way that a general-purpose compressor achieves improved
+compression ratios.
+
+% In the realm of protein structure prediction, it has been
+% observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous
+% range of
+% possible dihedral angles, protein structures inhabit a smaller subspace, defined
+% by a narrower range of angles, referred to as
+% rotamers~\cite{dunbrack1993backbone}.
+% The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation
+% for the
+% purpose of sidechain packing (finding the lowest-energy conformation for the
+% sidechain of each amino acid), using tree width to reduce the search space and
+% simplify dependencies.
 
@@
 
-\begin{table}
-\caption{Metric entropy ratio (ratio of clusters to entries in database) and
-fractal dimension at typical search radii for four biological data sets.
-NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
-\begin{tabular}{ccc}
-\hline
-Data set & Metric entropy ratio & Fractal dimension \\        
-\hline
-Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
-\hline
-Protein sequences (NCBI NR) & 5:1 & 1.6\\
-\hline
-Protein structure (PDB) & 10:1 & 2.5\\
-\hline
-Chemical structure (PubChem) & 11:1 & 0.2\\
-\hline
-\end{tabular}
-\end{table}
-
 % Bibliography
-% \bibliographystyle{ACM-Reference-Format-Journals}
-\bibliographystyle{abbrv}
+\bibliographystyle{ACM-Reference-Format-Journals}
+% \bibliographystyle{abbrv}
 \bibliography{main}

### 2015-07-26 main.tex
Structure-of-data section, and Bonnie edits

@@
 researchers~\cite{marx2013biology}.
-Currently, many omics applications require us to store, access, and analyze
+Currently, many biological 'omics` applications require us to store, access, and analyze
 large libraries of data.
@@
 Amazon Web Services are also commonly used for computational biology 
-research~\cite{schatz2010cloud}.
+research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
 However, while cloud computing frees researchers from maintaining their own
@@
 not needed around the clock, it is no panacea.
-First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
+First and foremost, cloud computing does not truly address the problem posed by
+a corpus of data that is growing faster than Moore's law, because the computer
+systems that make up those cloud datacenters are still themselves bound by
+improvements in semiconductor technology; they still follow the trajectory
+suggested by Moore's law.
+Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
 in West Africa, analysis resources are needed at often-remote field sites.
@@
 Thus, accessing cloud resources for analytics may not be possible.
-Moreover, cloud computing does not truly address the problem posed by
-a corpus of data that is growing faster than Moore's law, because the computer
-systems that make up those cloud datacenters are still themselves bound by
-improvements in semiconductor technology; they still follow the trajectory
-suggested by Moore's law.
 
-Computer scientists routinely exploit peculiarities in the structure of data in
+Computer scientists routinely exploit the structure of various data in
 order to reduce time or space complexity.
-In computational biology, this approach has served researchers well.
+In computational biology, this approach has implicitly served researchers well.
 Now-classical approaches such as principal component analysis (PCA) reduce the 
@@
 Transform (BWT) also take advantage of aspects of sequence 
-data~\cite{berger2013computational}.
-This review focuses on algorithmic advances for dealing with the growth in 
-biological data; algorithms for gaining novel biological insights are not its 
+data~\cite{berger2013computational} to speed up computation and save storage.
+This review focuses on cutting-edge algorithmic advances for dealing with the growth in 
+biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
 focus.
@@
 Sequence data are obtained in several ways, including mass spectrometry (which
-can determine protein sequence) and RNA-seq (which can determine RNA sequence, 
-and by inference the protein sequence into which it might translate).
+can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
+and by inference the expression  of the gene it might translate).
 However, the greatest volume of sequence data derive from next-generation 
 sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
-At the dawn of the genomic era, Sanger sequencing was the most widely used 
+At the dawn of the genomic era, Sanger sequencing was the most widely-used 
 method.
@@
 Instead, DNA molecules are chopped into many small fragments, or 
-\emph{reads} (Figure ~\ref{fig:shotgun}), 
+\emph{reads} (Figure~\ref{fig:shotgun}), 
 which are read concurrently, and which must be reassembled in the correct order 
 if an entire genome is to be read.
-Reads most typically range from 50 to 200 bases long, though longer reads are
-available with some technologies.
+Current reads most typically range from 50 to 200 bases long, though longer reads are
+available with some technologies (e.g., PacBio).
 This so-called ``shotgun sequencing'' is analogous to the problem of 
@@
 file.
+
+Another important type of biological data is gene expression, or the quantity of
+a particular gene product (usually a protein) present in the cell.
+Gene expression data is useful for relating genotype to phenotype, and is 
+important in studying diseases including cancer.
+Usually obtained through microarrays or RNA-seq technologies, expression data is
+quantitative, as each gene from a sample is associated with a numeric expression
+level, and high-dimensional, as many thousands of genes may be analyzed at a
+time.
 
@@
 represented as labeled graphs.
-In such a graph, the vertices are atoms (labeled with chemical element) and
-the edges are bonds between atoms (labeled with bond type, such as single or
+In such a graph, the vertices are atoms (labeled with a chemical element) and
+the edges are bonds between atoms (labeled with a bond type, such as single or
 double).
@@
 
-\section{Challenges in biological data}
+Networks are frequently used to represent biological data, such as the genetic
+and physical interactions among proteins, such as those in metabolic 
+pathways~\cite{berger2013computational}.
+
+\section{Challenges with biological data}
 
@@
 When a reference sequence is available, NGS reads must be \emph{mapped} onto
-this reference (Figure ~\ref{fig:ngs-pipeline}).
-This allows the differences between the newly-sequenced genome and the reference
+this reference (Figure~\ref{fig:ngs-pipeline}).
+Mapping allows the differences between the newly-sequenced genome and the reference
 to be analyzed; these differences may include single-nucleotide polymorphisms
 (SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
-insertions 
-or deletions, or larger-scale changes in the genome.
+insertions or deletions, or larger-scale changes in the genome.
 While reference-based read mapping is a fundamentally simpler problem than
@@
 range from millions (for bacteria) to billions (for mammals).
-
-Standard techniques for reference-based read mapping rely on algorithmic 
-approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
-which provides efficient string 
-compression through a reversible transformation, while the FM-index data 
-structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
-based on the BWT, which provides efficient storage as well as fast search.
-BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
-Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
-efficient mapping of NGS reads.
-The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
-coupled with dynamic programming in a compressed representation, in order to 
-prune the search space when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
-the space of possible matches, likewise pruning the search space.
-Even with these approaches, read mapping remains a significant bottleneck in
-genomic research~\cite{hach2010mrsfast}.
 
@@
 matches in databases of known gene or protein sequences.
-
 Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
-the standard too for performing homology search on databases of nucleotide and
+the standard tool for performing homology (similarity) search on databases of nucleotide and
 protein sequences.
@@
 database hit.
-However, BLAST's running time is linear in the size of the database being
+However, BLAST's running time scales linearly with the size of the database being
 searched, which is problematic as sequence databases continue to grow at a
@@
 respectively.
+These genomes are enabling cross-species inference, and thus insights into 
+evolution~\cite{berger2013computational}.
+
 On a potentially even larger scale is the growth of metagenomic data.
@@
 obesity~\cite{greenblum2012metagenomic}.
-
 Metagenomics fundamentally asks what organisms are present, and (in the case
@@
 a whole.
-One approach to this is to attempt to map NGS reads from a metagenomic sample
+One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
 onto a set of reference genomes that are expected to be present.
@@
 reference genomes, compounding the computational requirements.
-A second approach is to perform homology search on a protein sequence database;
+A second way is to perform homology search on a protein sequence database;
 exact or nearly-exact matches imply the presence of a species, while more 
@@
 search for them in a protein database.
-Recent advances in metagenomic search tools have relied on two advances over
-BLASTX: indexing and alphabet reduction.
-RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
-collision-free hash table.
-The alphabet reduction, as it is reversible, can be thought of as a form of
-lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
-alphabet, with offsets stored to recover the original sequence in the full
-alphabet.
-The hash-table provides an efficient index of the database to be searched.
-DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
-``shaped seeds'' instead of simple $k$-mer seeds to index the database.
-DIAMOND demonstrates search performance three to four orders of magnitude faster
-than BLASTX, but it is still linear in the size of the database being searched.
+The difficulty is that the datasets required to shine any light on these 
+questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
+complex than 16S metagenomic studies that merely classify the bacteria based on 
+their divergence with respect to a specific gene, resulting in identification 
+problems for certain bacterial, as well as viral, species and 
+genera~\cite{janda200716s}.
+
+While gene expression data lends itself to cluster analysis and probabilistic 
+approaches, the high dimensionality and noise in the data presents significant 
+challenges~\cite{nguyen2002tumor}.
+Principal Component Analysis has shown promise in reducing the dimensionality
+of gene expression 
+data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
 
@@
 
-Another important type of biological data is gene expression, or the quantity of
-a particular gene product (usually a protein) present in the cell.
-Gene expression data is useful for relating genotype to phenotype, and is 
-important in studying diseases including cancer.
-Usually obtained through microarrays or RNA-seq technologies, expression data is
-quantitative, as each gene from a sample is associated with a numeric expression
-level, and high-dimensional, as many thousands of genes may be analyzed at a
-time.
-While expression data lends itself to cluster analysis and probabilistic 
-approaches, the high dimensionality presents challenges~\cite{nguyen2002tumor}.
-Principal Component Analysis has shown promise in reducing the dimensionality
-of gene expression data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
-Some recent work has explored other ways to exploit the high-dimensional 
+The study of drugs and their targets based on chemical structure and function
+is known as chemogenomics~\cite{bredel2004chemogenomics}.
+In the fields of drug discovery and drug repurposing, the prediction 
+of biologically active compounds is an important task. 
+Computational high-throughput screening eliminates many compounds from 
+laborious wet-lab consideration, but even computational screening can be 
+time-consuming.
+Notably, large chemical compound databases, such as PubCHEM,
+cannnot be searched on a laptop computer with standard tools (e.g., SMSD).
+
+With biological networks, challenges lie in integrating such data in order to 
+deduce gene function and disease pathways.
+Several approaches have taken advantage of the particular topologies of 
+biological
+networks, such as diffusion-based approaches, which explore the topology of
+networks through random walks.
+In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
+interact with
+many partners; these interactions do not imply shared function as strongly as
+non-hub interactions; such information needs to be teased apart.
+As new functional annotations and protein-protein interactions are discovered,
+the number and size of these networks grows, making related problems more 
+intractable~\cite{berger2013computational}.
+
+The continued ability to search and analyze these growing data sets hinges on
+clever algorithms that take advantage of the structure of, and redundancy 
+present in, the data.
+Indeed, these growing data sets ``threaten to make the arising problems 
+computationally infeasible.''~\cite{berger2013computational}
+
+\section{State-of-the-art approaches to meet these challenges}
+
+Techniques for reference-based read mapping typically rely on algorithmic 
+approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
+which provides efficient string 
+compression through a reversible transformation, while the FM-index data 
+structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
+based on the BWT, which provides efficient storage as well as fast search.
+BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
+Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
+efficient mapping of NGS reads.
+The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
+coupled with dynamic programming in a compressed representation, in order to 
+prune the search space when mapping reads to a reference genome.
+Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
+the space of possible matches, likewise pruning the search space.
+Even with these approaches, read mapping remains a significant bottleneck in
+genomic research~\cite{hach2010mrsfast}.
+
+Recent advances in metagenomic search tools have relied on two advances over
+BLASTX: indexing and alphabet reduction.
+RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
+collision-free hash table.
+The alphabet reduction, as it is reversible, can be thought of as a form of
+lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
+alphabet, with offsets stored to recover the original sequence in the full
+alphabet.
+The hash-table provides an efficient index of the database to be searched.
+DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
+``shaped seeds'' instead of simple $k$-mer seeds to index the database.
+DIAMOND demonstrates search performance three to four orders of magnitude faster
+than BLASTX, but it is still linear in the size of the database being searched.
+
+Recent work on gene expession has explored additional ways to exploit the high-dimensional 
 structure of the data.
@@
 predict relationships between genes.
-
 Another recent and novel approach to exploiting the structure of gene expression
@@
 
-Networks are frequently used to represent biological data, such as the genetic
-and physical interactions among proteins, such as those in metabolic pathways.
-Several approaches have taken advantage of the particular topologies of these
-networks, such as diffusion-based approaches, which explore the topology of
-networks through random walks.
-IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn} perform global multiple network 
-alignment based on network topology and conserved biological function.
+An area of great interest in integrating network data is in global multiple network 
+alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
 IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
@@
 PageRank-Nibble algorithm~\cite{andersen2006local}.
-In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
-interact with
-many partners; these interactions do not imply shared function as strongly as
-non-hub interactions.
-Diffusion state distance (DSD)~\cite{cao2013going} is another diffusion-based approach
+Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
 which defines a distance metric based on random walks.
@@
 functions when applied to standard approaches such as neighbor voting for function annotation.
-
-As new functional annotations and protein-protein interactions are discovered,
-the number and size of these networks grows.
 Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
@@
 prediction accuracy.
-
-The continued ability to search and analyze these growing data sets hinges on
-clever algorithms that take advantage of the structure of, and redundancy 
-present in, the data.
-Indeed, these growing data sets ``threaten to make the arising problems 
-computationally infeasible.''~\cite{berger2013computational}
 
@@
 low \emph{metric entropy}~\cite{yu2015entropy}.
-That is, for a given cluster radius $r_c$ and a database $D$, the number 
-$N_{r_c} (D)$ of clusters needed to cover $D$ is relatively small compared to
+That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
+of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
+entropy, which is relatively small compared to
 $|D|$, the number of entries in the database.
@@
 by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
-That is, for some range of radii $r_1$ and $r_2$,
+That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
+in the database $D$, the fractal dimension $d$ is
+$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
+number of points within $r_1$ and $r_2$ respectively.
 
-discrete?
+Cluster-based search, as exemplified by ``compressive omics,'' can perform
+approximate search within a radius $r$ of a query $q$ on a database $D$ with 
+fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
+proportional to
+\begin{gather*}
+    O\Bigg(
+    \underbrace{k}_{\textrm{metric entropy}} +
+    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
+    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
+     \Bigg) .
+\end{gather*}
 
-By combining redundancy with a low fractal 
-dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
-scale even as the data continue to grow.
+Given this formalization, the ratio $\frac{|D|}{k}$ provides an estimate of the 
+speed-up factor for the coarse search component compared to a full linear 
+search.
+The time-complexity of the fine search is exponential in the fractal dimension
+$d$, which can be estimated globally by sampling the local fractal dimension 
+over a dataset.
+Table~\ref{fractal} provides the fractal dimension $d$ sampled at typical query
+radii, as well as the ratio $\frac{|D|}{k}$, for nucleotide sequence, protein 
+sequence, protein structure, and chemical compound databases.
 
-wherever physical or empirical laws have constrained data to 
-a subspace of low entropy and fractal dimension.
-
-Specifically, the ratio $\frac{|D|}{k}$ provides an estimate of the acceleration factor
-for just the coarse search component compared to a full linear search.
-Local fractal dimension around a data point can be computed by determining the
-number of other data points within two radii $r_1$ and $r_2$ of that point;
-given those point counts ($n_1$ and $n_2$, respectively), fractal dimension $d$
-is simply $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$.
-Sampling this property over a dataset can provide a global average fractal 
-dimension.
-When we search a larger radius around a query, the number of points we encounter grows exponentially with the fractal dimension;
-low fractal dimension implies that this growth will not obviate the gains provided by an entropy-scaling data structure.
-
-An entropy-scaling search tree is also related to a metric ball tree~\cite{uhlmann1991satisfying}, although with different time complexity.
-Querying a metric ball tree requires $O(\log n)$ time, assuming the relatively uniform distribution of data points in a metric space.
-This is different from the nonuniform distribution under which entropy-scaling
-search behaves well.
+Biological data sets exhibit redundancy, and are constrained to subspaces by
+physical laws.
+This combination results in low fractal dimension and low metric entropy 
+relative to the size of the data set, which suggests that ``compressive omics''
+will provide the ability for computation to scale with massively growing data.
 
 \begin{table}
+\label{fractal}
 \caption{Metric entropy ratio (ratio of clusters to entries in database) and

### 2015-07-26 main.tex
Bonnie edits and l-mer discussion

@@
 On the one hand, the scale and scope of data should allow new insights into
-genetic and infections diseases, cancer, basic biology, and even human migration
+genetic and infectious diseases, cancer, basic biology, and even human migration
 patterns.
@@
 As another example, clever indexing techniques such as the Burroughs-Wheeler 
-Transform (BWT) also take advantage of aspects of sequence 
+Transform (BWT) take advantage of aspects of sequence 
 data~\cite{berger2013computational} to speed up computation and save storage.
@@
 This structure ultimately determines the protein's function within the cell.
-Certain kinds of RNA also function, sometimes with proteins, as cellular 
+Certain kinds of RNA also importantly function, sometimes with proteins, as cellular 
 machines.
@@
 molecule end-to-end as one could a file from a hard drive.
-Instead, DNA molecules are chopped into many small fragments, or 
-\emph{reads} (Figure~\ref{fig:shotgun}), 
-which are read concurrently, and which must be reassembled in the correct order 
+Instead, DNA molecules are chopped into many small fragments;
+from these we can measure 
+\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
+These reads are read concurrently and must be reassembled in the correct order 
 if an entire genome is to be read.
@@
 Another important type of biological data is gene expression, or the quantity of
-a particular gene product (usually a protein) present in the cell.
+a particular gene product (usually a protein and more recently RNA) present in the cell.
 Gene expression data is useful for relating genotype to phenotype, and is 
@@
 
-Protein structure, primarily determined through X-ray crystallography or
+Protein structure data, primarily determined through X-ray crystallography or
 nuclear magnetic resonance (NMR), describes the coordinates of each atom, in 
@@
 Networks are frequently used to represent biological data, such as the genetic
-and physical interactions among proteins, such as those in metabolic 
+and physical interactions among proteins, as well as those in metabolic 
 pathways~\cite{berger2013computational}.
@@
 Genome Project.
-When a reference sequence is available, NGS reads must be \emph{mapped} onto
+When a reference sequence is available, NGS reads can be \emph{mapped} onto
 this reference (Figure~\ref{fig:ngs-pipeline}).
@@
 at the heart of current research in computational biology.
-Since some organisms (known as \emph{model organisms}) are better-studied than
+Since some organisms (known as \emph{model organisms}) are better studied than
 others, and evolution is known to conserve sequence, structure, and function,
@@
 respectively.
-These genomes are enabling cross-species inference, and thus insights into 
+These genomes are enabling cross-species inference, for example about genes and regulatory regions, and thus insights into function and
 evolution~\cite{berger2013computational}.
+But as we saw earlier, their sheer growth is problematic for storage, access and analysis.
 
@@
 obesity~\cite{greenblum2012metagenomic}.
-Metagenomics fundamentally asks what organisms are present, and (in the case
-of a microbiome such as the gut) what metabolic functions it can accomplish as
+Metagenomics fundamentally asks what organisms are present, and, in the case
+of a microbiome such as the gut, what metabolic functions it can accomplish as
 a whole.
@@
 complex than 16S metagenomic studies that merely classify the bacteria based on 
-their divergence with respect to a specific gene, resulting in identification 
-problems for certain bacterial, as well as viral, species and 
+their divergence with respect to a specific gene.
+The massive data results in major identification 
+challenges for certain bacterial, as well as viral, species and 
 genera~\cite{janda200716s}.
@@
 data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
+Such data and its challenges has been the focus of other 
+articles~\cite{berger2013computational}, and thus will be only lightly touched 
+upon here.
 
@@
 indicates the abundance of particular, short structural motifs within a protein.
+%What are the challenges with this data?
 
@@
 
-The continued ability to search and analyze these growing data sets hinges on
+The continued ability to store, search, and analyze these growing data sets hinges on
 clever algorithms that take advantage of the structure of, and redundancy 
@@
 even when seeds are clustered. 
-Through its k-mer based read compression model, CORA is able to accelerate and 
+Through its $l$-mer based read compression model, CORA is able to accelerate 
+and 
 achieve asymptotically sublinear scaling for both the seed-matching and 
@@
 the read-mapping computation.
-
-NEED TO DISCUSS $l$-mers here.
+Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
+not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
+CORA uses much longer $k$-mers, and links each one to its neighbors within
+a small Hamming or Levenshtein distance.
+The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
 
@@
 proportional to
-\begin{gather*}
+\[
     O\Bigg(
@@
      \Bigg) .
-\end{gather*}
+\]
 

### 2015-07-26 main.tex
minor tweaks

@@
 to generating biological data; next-generation sequencing, mass spectroscopy,
-microarrays, nuclear magnetic resonance, and other high-throughput approaches
-have led to an explosion of data.
-This explosion of data is a mixed blessing.
+microarrays, nuclear magnetic resonance spectroscopy, and other high-throughput 
+approaches have led to an explosion of data.
+However, this explosion of data is a mixed blessing.
 On the one hand, the scale and scope of data should allow new insights into
@@
 As a result, the amount of genomic data available to 
-researchers is increasing by a factor of 10 every year.
+researchers is increasing by a factor of ten every year.
 
@@
 Protein structure data, primarily determined through X-ray crystallography or
-nuclear magnetic resonance (NMR), describes the coordinates of each atom, in 
-Angstroms, of a protein in three-dimensional space.
+nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
+each atom, in Angstroms, of a protein in three-dimensional space.
 This information is important to understanding how a protein functions, because

### 2015-07-26 main.tex
minor edits
a few more tweaks

@@
 researchers~\cite{marx2013biology}.
-Currently, many biological 'omics` applications require us to store, access, and analyze
-large libraries of data.
+Currently, many biological 'omics` applications require us to store, access, 
+and analyze large libraries of data.
 One approach to solving these challenges is to embrace cloud computing.
-Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
+Google, Inc. and the Broad Institute have collaborated to bring the GATK 
+(Genome 
 Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
@@
 data centers, and provides cost-saving benefits when computing resources are
-not needed around the clock, it is no panacea.
+not needed continuously, it is no panacea.
 First and foremost, cloud computing does not truly address the problem posed by
@@
 acids, known as a protein.
-This protein folds into a complex, low-energy structure, and functions as
-a cellular machine itself; the DNA sequence determines the amino acid sequence,
+This protein folds into a complex, low-energy structure, which functions as
+a cellular machine; the DNA sequence determines the amino acid sequence,
 which in turn determines the folded structure of the protein.
 This structure ultimately determines the protein's function within the cell.
-Certain kinds of RNA also importantly function, sometimes with proteins, as cellular 
+Certain kinds of RNA also function, sometimes with proteins, as cellular 
 machines.
@@
 Researchers also study the resulting functions of RNA and proteins, in what
-abundance they occur (this is known as gene expression), and how they interact
+abundance they occur (gene expression), and how they interact
 with each other (interaction networks).
@@
 can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
-and by inference the expression  of the gene it might translate).
+and by inference the expression of the gene to which it might translate).
 However, the greatest volume of sequence data derive from next-generation 
-sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
+sequencing (NGS) technologies, and comprise DNA sequences.
 At the dawn of the genomic era, Sanger sequencing was the most widely-used 
@@
 beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
-greater throughput due to massive parallelism, low cost, and simple preparation
-of samples.
+greater throughput due to massive parallelism, low cost, and simple sample
+preparation.
 Importantly to this Review, Illumina sequencing, as well as other approaches

### 2015-07-27 main.tex
Bonnie's edits incorporated

@@
 time.
+Resources such as NCBI’s Gene Expression Omnibus (GEO), which pool 
+together many diverse expression studies, now allow researchers to analyze 
+thousands of samples generated by others to obtain meaningful biological 
+insights. 
+More importantly, many of these biological insights can only be gleaned when 
+viewed in the context of hundreds or thousands of gene expression samples.
 
@@
 range from millions (for bacteria) to billions (for mammals).
+As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.
 
@@
 
-Given a sequenced genome, the next natural questions are what genes (genomic
-regions that code for proteins) are present, what structure each resulting
-protein takes, and what biological function it performs.
-Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
-beyond the scope of this Review.
-However, determining evolutionary relationships, structure, and function are
-at the heart of current research in computational biology.
-Since some organisms (known as \emph{model organisms}) are better studied than
-others, and evolution is known to conserve sequence, structure, and function,
-a powerful approach to determine these attributes is to search for similar
-sequences about which more is known.
-This so-called \emph{homology search} entails searching for \emph{approximate}
-matches in databases of known gene or protein sequences.
-Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
-the standard tool for performing homology (similarity) search on databases of nucleotide and
-protein sequences.
-BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
-that might lead to longer matches, and greedily extends them, ultimately 
-producing a \emph{sequence alignment} between a query and each potential 
-database hit.
-However, BLAST's running time scales linearly with the size of the database being
-searched, which is problematic as sequence databases continue to grow at a
-faster rate than Moore's law.
-
 Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
@@
 evolution~\cite{berger2013computational}.
-But as we saw earlier, their sheer growth is problematic for storage, access and analysis.
+Again, the sheer enormity of sequencing data is problematic for storage, access and analysis.
 
@@
 
+Given a sequenced genome, the next natural questions are what genes (genomic
+regions that code for proteins) are present, what structure each resulting
+protein takes, and what biological function it performs.
+Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
+beyond the scope of this Review.
+However, determining evolutionary relationships, structure, and function are
+at the heart of current research in computational biology.
+Since some organisms (known as \emph{model organisms}) are better studied than
+others, and evolution is known to conserve sequence, structure, and function,
+a powerful approach to determine these attributes is to search for similar
+sequences about which more is known. 
+This so-called \emph{homology search} entails searching for \emph{approximate}
+matches in databases of known gene or protein sequences.
+The homology search problem was previously believed to be solved;
+Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
+the standard tool for performing homology (similarity) search on databases of nucleotide and
+protein sequences.
+BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
+that might lead to longer matches, and greedily extends them, ultimately 
+producing a \emph{sequence alignment} between a query and each potential 
+database hit.
+However, BLAST's running time scales linearly with the size of the database being
+searched, which is problematic as sequence databases continue to grow at a
+faster rate than Moore's law.
+
 While gene expression data lends itself to cluster analysis and probabilistic 
@@
 data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
-Such data and its challenges has been the focus of other 
+Such data and its challenges have been the focus of other 
 articles~\cite{berger2013computational}, and thus will be only lightly touched 
@@
 Matt~\cite{menke2008matt}.
+Due to the complexity of protein structures, these programs generally take 
+significant amounts of time,
+especially to align multiple structures.
+For example, the DALI webserver can take as much as an hour to return results
+for a single query.
 FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
-approximating structural alignments, which can be computationally demanding,
+approximating structural alignments
 by instead comparing the ``bag-of-words'' from each structure.
@@
 indicates the abundance of particular, short structural motifs within a protein.
-%What are the challenges with this data?
+%Does Fragbag have challenges we should mention?
+% NMD: not really. That we speed it up is evidence of the orthogonality of
+% our approach
 
@@
 With biological networks, challenges lie in integrating such data in order to 
-deduce gene function and disease pathways.
+deduce gene function, as well as uncover disease and metabolic pathways.
 Several approaches have taken advantage of the particular topologies of 
@@
 non-hub interactions; such information needs to be teased apart.
-As new functional annotations and protein-protein interactions are discovered,
+As new functional annotations and interactions are discovered,
 the number and size of these networks grows, making related problems more 
@@
 BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
-Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
+Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
 efficient mapping of NGS reads.
 The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
-coupled with dynamic programming in a compressed representation, in order to 
-prune the search space when mapping reads to a reference genome.
+coupled with dynamic programming in a compressed representation of the 
+reference genome, in order to prune the search space 
+when mapping reads to a reference genome.
 Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
-the space of possible matches, likewise pruning the search space.
+the space of possible matches, likewise pruning the search space; however, the
+bulk of its runtime is spent on the extend phase.
 Even with these approaches, read mapping remains a significant bottleneck in
@@
 well-described by tetrahedra whose vertices were enriched for different tumor
-types and biological functions, from which~\cite{hart2015inferring} infer four distinct
-tasks.
+types and biological functions, from which four distinct tasks are 
+inferred~\cite{hart2015inferring}.
 
@@
 alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
-IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
+IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank} 
+but performs a random walk on the product graph of two networks;
 IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
@@
 Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
-which defines a distance metric based on random walks.
+which defines a distance metric between two nodes based on random walks.
 DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
@@
 network in such a way that function annotation can be performed in the reduced 
-space, significantly reducing running time while still providing competitive
+space, significantly decreasing running time while still providing competitive
 prediction accuracy.
+Using a compressed representation of metabolic networks, 
+Mongoose~\cite{chindelevitch2014exact} is able to 
+identify and unblock many blocked reactions in current models without loss of 
+accuracy.
 
@@
 Seeking to take advantage of the redundancy inherent in genomic sequence data, 
-\cite{loh2012compressive}
+Loh, Baym and Berger \cite{loh2012compressive}
 introduced \emph{compressive genomics}, an approach that relies on compressing
@@
 that represent unique data.
-Any coarse sequence that is within some threshold of the query is then expanded 
+Any representative sequence that is within some threshold of the query is then expanded 
 into all the similar sequences it represents; the fine search is then over this

### 2015-07-28 main.tex
treepoints figure, fix figure margins

@@
 \usepackage{subfig}
+\captionsetup[subfigure]{margin=5pt}
 
@@
 % Page heads
-\markboth{B. Berger et al.}{Title}
+\markboth{B. Berger et al.}{Computational biology in the 21st century: Algorithms that scale}
 
 % Title portion
-\title{Title}
+\title{Computational biology in the 21st century: Algorithms that scale}
 \author{BONNIE BERGER
@@
 \begin{abstract}
-abstract goes here
+Throughout all areas of data science, scientists are being confronted by a
+veritable explosion of available data. In many fields, this increase is
+exponential in nature, even outpacing Moore's and Kryder's laws on the
+respective doublings of transistors on a chip and long-term data storage
+density. As such, the challenges posed by the massive influx of data cannot
+be solved by waiting for faster and larger capacity computers, but require
+instead the development of data structures and representations that exploit
+and simplify complexity in the dataset. In this Review, we
+survey the biological data landscape, with a focus on how taking advantage
+of the structure of massive biological data enables the design of algorithms
+that scale sublinearly in large-scale genomics, personal genomics and
+chemogenomics.
 \end{abstract}
@@
 \begin{bottomstuff}
+This work is supported by the National Institutes of Health, under
+grant GM108348.
 % This work is supported by the National Science Foundation, under
@@
 \subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
-\includegraphics[width=3.0in]{assets/moore.png}
+\includegraphics[width=2.8in]{assets/moore.png}
 \label{fig:moore}
-}\\
+}
 \subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
-\includegraphics[width=3.0in]{assets/kryder.png}
+\includegraphics[width=2.8in]{assets/kryder.png}
 \label{fig:kryder}
@@
 Genome Project.
+
 When a reference sequence is available, NGS reads can be \emph{mapped} onto
 this reference (Figure~\ref{fig:ngs-pipeline}).
-Mapping allows the differences between the newly-sequenced genome and the reference
-to be analyzed; these differences may include single-nucleotide polymorphisms
+Mapping allows the differences between the newly-sequenced genome and the 
+reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
 (SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
 insertions or deletions, or larger-scale changes in the genome.
+Determining the differences between an individual genome and a reference is
+known as \emph{variant calling}.
 While reference-based read mapping is a fundamentally simpler problem than
@@
 range from millions (for bacteria) to billions (for mammals).
+As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
+(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types.
+Currently each sample requires seven hours to download. 
+Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
+The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
 As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.
@@
 \centering
-\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, or reads, and relies on high coverage to produce a statistically likely representation of a whole genome]{
-\includegraphics[width=3.0in]{assets/shotgun.png}
+\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
+\includegraphics[width=2.6in]{assets/shotgun.png}
 \label{fig:shotgun}
@@
 \subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
-\includegraphics[width=3.0in]{assets/snps.png}
+\includegraphics[width=2.6in]{assets/snps.png}
 \label{fig:snps}
@@
 This allows relationships between genes and diseases to be uncovered.]{
-\includegraphics[width=3.0in]{assets/ngs-pipeline.png}
+\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
 \label{fig:ngs-pipeline}
@@
 database hit.
-However, BLAST's running time scales linearly with the size of the database being
+However, BLAST's running-time scales linearly with the size of the database being
 searched, which is problematic as sequence databases continue to grow at a
@@
 indicates the abundance of particular, short structural motifs within a protein.
-%Does Fragbag have challenges we should mention?
-% NMD: not really. That we speed it up is evidence of the orthogonality of
-% our approach
 
-The study of drugs and their targets based on chemical structure and function
+The computational study of drugs and their targets based on chemical structure and function
 is known as chemogenomics~\cite{bredel2004chemogenomics}.
@@
 time-consuming.
-Notably, large chemical compound databases, such as PubCHEM,
-cannnot be searched on a laptop computer with standard tools (e.g., SMSD).
+Chemogenomics typically relies on comparing chemical graph structures to identify
+similar molecules and binding sites.
+Furthermore, comparing chemical graph structures typically involves computing
+the maximal common subgraph (MCS), an NP-hard problem.
+However, there are once again an increasing number of such chemical compounds
+to search; the NCBI's PubChem database has grown from 31 million compounds in
+January 2011 to 68 million in July, 2015.
 
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an ``approximate seed'' approach to indexing
 the space of possible matches, likewise pruning the search space; however, the
 bulk of its runtime is spent on the extend phase.
+State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
+only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
 Even with these approaches, read mapping remains a significant bottleneck in
@@
 
-Recent advances in metagenomic search tools have relied on two advances over
+Recent advances in metagenomic search tools have relied on two improvements over
 BLASTX: indexing and alphabet reduction.
@@
 
-An area of great interest in integrating network data is in global multiple network 
+The most widely-used chemogenomics search is the Small Molecule Subgraph Detector 
+(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
+on the size and complexity of the graphs in question.
+Notably, large chemical compound databases, such as PubCHEM,
+cannnot be searched on a laptop computer with current tools such as SMSD.
+
+An area of great interest in integrating biological network data is in global multiple network 
 alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
@@
 
-\section{Compressive omics}
+\section{Structure of biological data}\label{structure}
 
-Another approach is to make use of a completely different paradigm for the 
-structure of biological data.
-Much biological data is highly redundant; if a computation is performed on one
+Fortunately, biological data has unique structure \cite{yu2015entropy}, which we 
+will take advantage of to perform search that scales sublinearly in the size of the database~(Section~\ref{omics}).
+The first critical observation is that
+much biological data is highly redundant; if a computation is performed on one
 human genome, and a researcher wishes to perform the same computation on another
-human genome, most of the work has already been done.
+human genome, most of the work has already been done \cite{loh2012compressive}.
+When dealing with redundant data, clustering comes to mind.
+While cluster-based search is well-studied~\cite{jardine1971use}, conventional
+wisdom holds that it provides a constant factor speed-up over exhaustive search.
 
-Seeking to take advantage of the redundancy inherent in genomic sequence data, 
-Loh, Baym and Berger \cite{loh2012compressive}
-introduced \emph{compressive genomics}, an approach that relies on compressing
-data in such a way that the desired computation (such as BLAST search) can be
-performed in the compressed representation.
-This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
-Coarse search is performed only on the coarse, or representative, subsequences
-that represent unique data.
-Any representative sequence that is within some threshold of the query is then expanded 
-into all the similar sequences it represents; the fine search is then over this
-(typically small) subset of the original database.
-This approach provides run-time improvements to BLAST search on 
-nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
-data; moreover, these run-time improvements increase as databases grow.
-
-The CORA read mapper~\cite{yorukoglu2015compressive} applies an $l$-mer based 
-read-compression approach with a compressive indexing of the reference genome
-(referred to as a homology table).
-CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
-mappers including BWA, and Bowtie2) by allowing them to operate in a compressed
-space, and relying on a coarse and a fine phase.
-Short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
-MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
-schemes aim to accelerate only the seed-to-reference matching step.
-Thus, there is a subsequent seed-extension step, which is substantially more 
-costly and still needs to be performed for each read and mapping individually, 
-even when seeds are clustered. 
-Through its $l$-mer based read compression model, CORA is able to accelerate 
-and 
-achieve asymptotically sublinear scaling for both the seed-matching and 
-seed-extension steps within coarse-mapping, which comprises the major bulk of 
-the read-mapping computation.
-Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
-not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
-CORA uses much longer $k$-mers, and links each one to its neighbors within
-a small Hamming or Levenshtein distance.
-The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
-
-In the area of metagenomic search,
-the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
-compressive-acceleration approach of 
-caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
-largely orthogonal to alphabet-reduction and indexing approaches.
-MICA applies the compressive-acceleration framework to 
-DIAMOND~\cite{buchfink2014fast}, using DIAMOND
-for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
-``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
-caBLASTP does over BLASTP.
-
-The compressive acceleration approach of \cite{loh2012compressive} has been
-generalized and adapted to non-sequence spaces as well.
-Chemogenomics, the computational study of small-molecule compounds and their
-affect on phenotype, relies on comparing chemical graph structures to identify
-similar molecules and binding sites.
-However, there are once again an increasing number of such chemical compounds
-to search; the NCBI's PubChem database has grown from 31 million compounds in
-January 2011 to 68 million in July, 2015.
-Furthermore, comparing chemical graph structures typically involves computing
-the maximal common subgraph (MCS), an NP-hard problem.
-One approach to this type of search is the Small Molecule Subgraph Detector 
-(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
-on the size and complexity of the graphs in question.
-Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
-accelerates SMSD search by an average of 150x on the PubChem database.
-
-Again applying a compressive acceleration approach, 
-esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
-Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
-running time by an average of 10x.
-
-\section{Structure of biological data}
-
-Cluster-based search is well-studied~\cite{jardine1971use}, but conventional
-wisdom holds that it provides a constant factor speed-up over exhaustive search.
-However, \cite{daniels2013compressive} observed that caBLASTP demonstrated
-increased gains over BLASTP on newer, larger protein databases, suggesting that
-compressive acceleration takes advantage of ever-increasing redundancy in the
-data.
 Beyond redundancy, however, another attribute of large biological data sets
@@
 2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
-This is not simply due to undersampling; thanks to evolution, only those protein
+This surprisingly lower number is not simply due to undersampling; thanks to 
+evolution, only those protein
 sequences which exhibit useful biological function should survive, and most
@@
 Thus, not only do biological data exhibit redundancy, they also tend not to 
-inhabit the entire feasible space.
+inhabit anywhere near the entire feasible space (Figure~\ref{fig:clusters}).
 It seems that physical laws have constrained the data to a particular subspace
@@
 
-One key insight as to why the ``compressive omics'' approach allows computation
-that scales sublinearly in the size of the data is that such data sets exhibit
+\begin{figure}[htb!]
+\centering
+\subfloat[Cartoon depiction of points in an arbitrary high-dimensional space, 
+as might arise from genomes generated by mutation and selection during the 
+course of evolution. 
+Although high-dimensional locally, at the global scale of covering spheres, the 
+data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
+similarity search. Clusters cover the data points but do not cover unoccupied 
+regions of space.]{
+\includegraphics[width=2.8in]{assets/treepoints.pdf}
+\label{fig:clusters}
+}
+\subfloat[A query (green triangle) and two concentric search radii (red 
+circles) around it. Thanks to low fractal dimension, the large circle does
+not contain vastly more points than the small circle.]{
+\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
+\label{fig:fractal}
+}
+\caption[Data spaces.]{A cartoon illustration of a biological data space.}
+\label{fig:dataspace}
+\end{figure}
+
+One key insight related to redundancy
+%as to why the compressive omics approach allows computation 
+%that scales sublinearly in the size of the data 
+is that such data sets exhibit
 low \emph{metric entropy}~\cite{yu2015entropy}.
@@
 entropy, which is relatively small compared to
-$|D|$, the number of entries in the database.
+$|D|$, the number of entries in the database (Figure~\ref{fig:clusters}).
 In contrast, if the points were uniformly distributed about the Cartesian space,
@@
 
-A second key insight is that the biological data sets considered 
-by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
+A second key insight is that the biological data sets have low \emph{fractal dimension}~\cite{yu2015entropy}.
 That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
@@
 $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
-number of points within $r_1$ and $r_2$ respectively.
+number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:fractal}).
 
@@
 relative to the size of the data set, which suggests that ``compressive omics''
-will provide the ability for computation to scale with massively growing data.
+will provide the ability for computation to scale sublinearly with massively growing data.
 
@@
 
+\section{The age of compressive omics}\label{omics}
+
+We are entering the age of compressive omics, 
+which makes use of this completely different paradigm for the structure of biological data.
+Seeking to take advantage of the redundancy inherent in genomic sequence data, 
+Loh, Baym and Berger \cite{loh2012compressive}
+introduced \emph{compressive genomics}, an approach that relies on compressing
+data in such a way that the desired computation (such as BLAST search) can be
+performed in the compressed representation.
+Daniels et al. \cite{daniels2013compressive} further demonstrated that
+compressive acceleration takes advantage of increasing redundancy in the
+data over time.
+
+Compressive genomics is based on the concept of \emph{compressive acceleration},
+which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
+Coarse search is performed only on the coarse, or representative, subsequences
+that represent unique data.
+Any representative sequence that is within some threshold of the query is then expanded 
+into all the similar sequences it represents; the fine search is then over this
+(typically small) subset of the original database.
+This approach provides orders-of-magnitude run-time improvements to BLAST search on 
+nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
+data; moreover, these run-time improvements increase as databases grow.
+
+The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
+read-compression approach with a compressive indexing of the reference genome
+(referred to as a homology table).
+CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
+mappers including BWA or Bowtie2) by allowing them to operate in a compressed
+space, and relying on a coarse and a fine phase.
+In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
+MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
+schemes aim to accelerate only the seed-to-reference matching step.
+Thus, there is a subsequent seed-extension step, which is substantially more 
+costly and still needs to be performed for each read and mapping individually, 
+even when seeds are clustered. 
+Through its $l$-mer based read compression model, CORA is able to accelerate 
+and 
+achieve asymptotically sublinear scaling for both the seed-matching and 
+seed-extension steps within coarse-mapping, which comprises the major bulk of 
+the read-mapping computation.
+Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
+not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
+CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
+a small Hamming or Levenshtein distance.
+The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
+
+In the area of metagenomic search,
+the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
+compressive-acceleration approach of 
+caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
+largely orthogonal to alphabet-reduction and indexing approaches.
+MICA applies the compressive-acceleration framework to the
+state-of-the-art DIAMOND~\cite{buchfink2014fast}, using it
+for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
+``fine search'' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.
+
+Compressive genomics \cite{loh2012compressive} has been
+generalized and adapted to non-sequence spaces as well, and coined
+``compressive omics''~\cite{yu2015entropy}.
+One such example is chemogenomics.
+Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
+accelerates SMSD search by an average of 150x on the PubChem database.
+Another example is 
+esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
+Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
+running time by an average of 10x.
+
 \section{Storage compression}
@@
 Several approaches to compressing quality scores have been 
-presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
+presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality,yu2014traversing}.
 Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
 compression, by bounding the likelihood that a quality score is informative.
+Importantly, Quartz modifies the FASTQ file of the reads and quality scores, and thus can be
+readily incorporated into existing sequence analysis pipelines.
 Surprisingly, Quartz not only demonstrates improved compression over competing
@@
 representation comprises only its unique differences from the bucket label.
-This allows a general-purpose compressor to achieve tight compression.
+This approach allows a general-purpose compressor to achieve tight compression.
 SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
@@
 ball tree, first described in the database community over twenty years 
-ago~\cite{uhlmann1991satisfying}.
-Other ideas from image processing, computational geometry, and other areas 
-outside of biology are likely to bear fruit.
-It is also likely that algorithmic ideas developed within computational biology
+ago~\cite{uhlmann1991satisfying};
+however, the latter does not allow one to analyze performance guarantees, 
+in terms of metric entropy and fractal dimension.
+Other ideas from image processing, computational geometry, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
+outside of biology are likely to bear fruit~\cite{indyk1998approximate}. 
+In turn, it is also likely that algorithmic ideas developed within computational biology
 will become useful in other fields experiencing a data deluge, such as astronomy
@@
 
+\section{Acknowledgments}
+
 % Bibliography

### 2015-07-28 main.tex
ACM fails single-point-of-truth for titles.
Mass spectroscopy ==> mass spectrometry
Spectroscopy is an old and now discouraged term, though correct.

@@
 \acmformat{Bonnie Berger, Noah M. Daniels,
-and Y. William Yu, 2015. Title.}
+and Y. William Yu, 2015. Computational biology in the 21st century: Algorithms that scale.}
 % At a minimum you need to supply the author names, year and a title.
@@
 The past two decades have seen unprecedented technological progress with regard
-to generating biological data; next-generation sequencing, mass spectroscopy,
+to generating biological data; next-generation sequencing, mass spectrometry,
 microarrays, nuclear magnetic resonance spectroscopy, and other high-throughput 

### 2015-07-28 main.tex
Fixed omics smartquotes
Lots of glue added to types and sources of biological data, and some minor edits

@@
 researchers~\cite{marx2013biology}.
-Currently, many biological 'omics` applications require us to store, access, 
+Currently, many biological `omics' applications require us to store, access, 
 and analyze large libraries of data.
@@
 However, while cloud computing frees researchers from maintaining their own
-data centers, and provides cost-saving benefits when computing resources are
+data centers and provides cost-saving benefits when computing resources are
 not needed continuously, it is no panacea.
-First and foremost, cloud computing does not truly address the problem posed by
-a corpus of data that is growing faster than Moore's law, because the computer
-systems that make up those cloud datacenters are still themselves bound by
-improvements in semiconductor technology; they still follow the trajectory
-suggested by Moore's law.
+First and foremost, the computer systems that make up those cloud datacenters
+are themselves bound by improvements in semiconductor technology and Moore's
+law. Thus, cloud computing does not truly address the problem posed by the
+faster-than-Moore's law exponential growth in omics data.
 Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
@@
 Transform (BWT) take advantage of aspects of sequence 
-data~\cite{berger2013computational} to speed up computation and save storage.
+structure~\cite{berger2013computational} to speed up computation and save storage.
 This review focuses on cutting-edge algorithmic advances for dealing with the growth in 
@@
 In the central dogma of molecular biology, DNA is transcribed into RNA, which
-is translated by the ribosome into a polypeptide chain, a sequence of amino 
-acids, known as a protein.
-This protein folds into a complex, low-energy structure, which functions as
-a cellular machine; the DNA sequence determines the amino acid sequence,
-which in turn determines the folded structure of the protein.
-This structure ultimately determines the protein's function within the cell.
+is translated by the ribosome into polypeptide chains, sequences of amino 
+acids which singly or in complexes are known as proteins.
+Proteins fold into complex, low-energy structures, which function as
+cellular machines; the DNA sequence determines the amino acid sequence,
+which in turn determines the folded structure of a protein.
+This structure ultimately determines a protein's function within the cell.
 Certain kinds of RNA also function, sometimes with proteins, as cellular 
@@
 abundance they occur (gene expression), and how they interact
-with each other (interaction networks).
+with both each other (interaction networks) and small chemical compound agonists.
 
 Most but not all biological data take the form of \emph{sequence} data, either 
-nucleotide sequences (nominally a four-letter alphabet representing the four 
-DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
+nucleotide sequences (using a four-letter alphabet representing the four 
+DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
 representing the twenty standard amino acids).
 Sequence data are obtained in several ways, including mass spectrometry (which
-can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
-and by inference the expression of the gene to which it might translate).
-However, the greatest volume of sequence data derive from next-generation 
-sequencing (NGS) technologies, and comprise DNA sequences.
+can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
+allowing scientists to also infer the expression of the gene to which it might translate).
+However, the greatest volume of sequence data available are genomic in nature and 
+come from next-generation sequencing (NGS) technologies.
+
 At the dawn of the genomic era, Sanger sequencing was the most widely-used 
-method.
-Slow and labor intensive but well-tested, Sanger sequencing is still used to 
-validate NGS results.
-More recently, Sanger sequencing has largely been supplanted by NGS approaches,
+method for reading a genome.
+Though slow and labor intensive, Sanger sequencing has the advantage of being
+well-tested and is still used to validate NGS results.
+More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
 beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
@@
 preparation.
-Importantly to this Review, Illumina sequencing, as well as other approaches
-such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
-molecule end-to-end as one could a file from a hard drive.
+
+Importantly to this Review, Illumina sequencing and other NGS approaches
+such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
+molecule end-to-end as one could read through a bound book.
 Instead, DNA molecules are chopped into many small fragments;
-from these we can measure 
+from these fragments we generate 
 \emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
 These reads are read concurrently and must be reassembled in the correct order 
-if an entire genome is to be read.
-Current reads most typically range from 50 to 200 bases long, though longer reads are
+to piece together an entire genome.
+Current reads typically range from 50 to 200 bases long, though longer reads are
 available with some technologies (e.g., PacBio).
 This so-called ``shotgun sequencing'' is analogous to the problem of 
-reconstructing a large file from small, 
-randomly-ordered reads, or a shredded paper document.
+reconstructing a book with all its pages torn out.
 However, these NGS reads overlap, and they are many in number, so 
 reconstructing an entire genome is feasible.
-No sequencing technology is completely infallible, so sequencing machines also
+
+Additionally, no sequencing technology is completely infallible, so sequencing machines also
 provide a \emph{quality score} (or measure of the confidence in the DNA base
@@
 
-Another important type of biological data is gene expression, or the quantity of
-a particular gene product (usually a protein and more recently RNA) present in the cell.
+While sequence is important, scientists are also interested in measuring gene expression, or the quantity of
+a particular gene product (usually a protein or more recently RNA) present in the cell.
 Gene expression data is useful for relating genotype to phenotype, and is 
-important in studying diseases including cancer.
+important in studying many diseases.
 Usually obtained through microarrays or RNA-seq technologies, expression data is
@@
 
+As mentioned earlier, function follows form, so in addition to
+sequence and expression, structure plays an important role in biological data science.
 Protein structure data, primarily determined through X-ray crystallography or
@@
 This information is important to understanding how a protein functions, because
-proteins interact physically with other molecules.
+proteins interact physically with other molecules and are geometrically constrained.
 Protein structure data are commonly stored in the Protein Data Bank (PDB); 
@@
 
-Chemical compounds represent another source of structural data, more general
-than proteins.
+Chemical compounds represent another source of structural data.
+Although generically much smaller than proteins, their structures are more general
+as they are not constrained in composition to amino acids.
 Determined by X-ray crystallography, NMR, electron microscopy and other 
-techniques, compounds---many of which are useful pharmaceuticals---can be
-represented as labeled graphs.
-In such a graph, the vertices are atoms (labeled with a chemical element) and
+techniques, chemical compounds---many of which are useful pharmaceuticals---can be
+represented as labeled graphs (or equivalently, adjacency matrices).
+In such graphs, the vertices are atoms (labeled with a chemical element) and
 the edges are bonds between atoms (labeled with a bond type, such as single or
 double).
-Chemical compounds are stored in databases such as PubChem, and can be described
-as adjacency matrices.
+Chemical compounds are stored in databases such as PubChem, and are often searched
+for by medicinal chemists as potential agonists for protein targets.
 
-Networks are frequently used to represent biological data, such as the genetic
+When examining more complex phenomena than single proteins or compounds, we
+often look to synthesize things together.
+To that end, we frequently use
+networks to represent biological data, such as the genetic
 and physical interactions among proteins, as well as those in metabolic 
 pathways~\cite{berger2013computational}.
+These relationships and interactions play crucial roles in a systems-level understanding of biology.
 

### 2015-07-28 main.tex
More minor edits

@@
 and analyze large libraries of data.
+% YWY: `omics' is the only single quote we use. Are we consistent?
 One approach to solving these challenges is to embrace cloud computing.
@@
 terabytes of reads must each be mapped onto the reference genome, which can 
-range from millions (for bacteria) to billions (for mammals).
+range from millions (for bacteria) to billions (for mammals) of base pairs.
 As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
@@
 Currently each sample requires seven hours to download. 
+% YWY: 7 hours for whom? Perhaps "even on an institutional connection"?
 Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
@@
 search for them in a protein database.
+% YWY: 16S isn't well introduced
 The difficulty is that the datasets required to shine any light on these 
 questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
-complex than 16S metagenomic studies that merely classify the bacteria based on 
+complex than the 16S metagenomic studies that merely classify the bacteria based on 
 their divergence with respect to a specific gene.
@@
 
-Given a sequenced genome, the next natural questions are what genes (genomic
+Given a sequenced genome, the next natural questions ask what genes (genomic
 regions that code for proteins) are present, what structure each resulting
@@
 beyond the scope of this Review.
-However, determining evolutionary relationships, structure, and function are
+However, determining evolutionary relationships, structure, and function is
 at the heart of current research in computational biology.
@@
 \section{Acknowledgments}
+Y.W.Y. is supported by a Hertz Foundation Fellowship.
 

### 2015-07-28 main.tex
More notes
new figure and minor tweaks

@@
 
+% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
 \begin{figure}[htb!]
@@
 }
-\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
+\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
+% YWY: My inclination is to write "(a) Moore's and (b) Kryder's laws". Possibly a matter of taste though, so do as you will.
 \label{fig:growth}
@@
 
+% YWY: "Most" is an extremely strong claim. I believe you, but I could see people
+% arguing (e.g. all of electronic health records could be considered "biological")
 Most but not all biological data take the form of \emph{sequence} data, either 
@@
 More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
-beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
+beginning with Illumina's `sequencing by synthesis,' which allows vastly 
 greater throughput due to massive parallelism, low cost, and simple sample
@@
 available with some technologies (e.g., PacBio).
-This so-called ``shotgun sequencing'' is analogous to the problem of 
+This so-called `shotgun sequencing' is analogous to the problem of 
 reconstructing a book with all its pages torn out.
@@
 
+% YWY: Take spell-check lines out of figure labels! (bp, Samtools)
 \begin{figure}[htb!]
 \centering
-\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
+\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
 \includegraphics[width=2.6in]{assets/shotgun.png}
@@
 }
-\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
+\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
 \includegraphics[width=2.6in]{assets/snps.png}
@@
 The difficulty is that the datasets required to shine any light on these 
-questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
-complex than the 16S metagenomic studies that merely classify the bacteria based on 
-their divergence with respect to a specific gene.
+questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
+complex than the 16S metagenomic studies that merely classify the bacteria 
+based on 
+their divergence with respect to a specific gene (namely, the 16S ribosomal 
+subunit).
 The massive data results in major identification 
@@
 protein sequences.
-BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
+BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
 that might lead to longer matches, and greedily extends them, ultimately 
@@
 approximating structural alignments
-by instead comparing the ``bag-of-words'' from each structure.
+by instead comparing the `bag-of-words' from each structure.
 Analogous to a term-frequency vector in information retrieval, this bag-of-words
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an ``approximate seed'' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an `approximate seed' approach to indexing
 the space of possible matches, likewise pruning the search space; however, the
@@
 DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
-``shaped seeds'' instead of simple $k$-mer seeds to index the database.
+`shaped seeds' instead of simple $k$-mer seeds to index the database.
 DIAMOND demonstrates search performance three to four orders of magnitude faster
@@
 This combination results in low fractal dimension and low metric entropy 
-relative to the size of the data set, which suggests that ``compressive omics''
+relative to the size of the data set, which suggests that `compressive omics'
 will provide the ability for computation to scale sublinearly with massively growing data.
@@
 fractal dimension at typical search radii for four biological data sets.
-NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
+NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
 \begin{tabular}{ccc}
@@
 Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
-not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
+not necessarily, a power of two) used as `seeds' for longer sequence matches.
 CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
@@
 state-of-the-art DIAMOND~\cite{buchfink2014fast}, using it
-for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
-``fine search'' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.
+for its `coarse search' phase and a user's choice of DIAMOND or BLASTX for its
+`fine search' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.
 
@@
 generalized and adapted to non-sequence spaces as well, and coined
-``compressive omics''~\cite{yu2015entropy}.
+`compressive omics'~\cite{yu2015entropy}.
 One such example is chemogenomics.
@@
 similar reads (those that share a common 
-substring) together into ``buckets'', allowing that common substring to now be 
+substring) together into `buckets', allowing that common substring to now be 
 removed and treated as the bucket label, so that each read in the compressed 
@@
 This approach allows a general-purpose compressor to achieve tight compression.
-SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
+SCALCE~\cite{hach2012scalce} also relies on a `boosting' scheme, reordering
 reads in such a way that a general-purpose compressor achieves improved
@@
 in terms of metric entropy and fractal dimension.
-Other ideas from image processing, computational geometry, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
-outside of biology are likely to bear fruit~\cite{indyk1998approximate}. 
+Other ideas from image processing, computational geometry~\cite{indyk1998approximate}, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
+outside of biology are likely to bear fruit. 
 In turn, it is also likely that algorithmic ideas developed within computational biology

### 2015-07-29 main.tex
Convert to ACM sigproc template.

@@
-% v2-acmsmall-sample.tex, dated March 6 2012
-% This is a sample file for ACM small trim journals
-%
-% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
-% (c) 2010 Association for Computing Machinery (ACM)
-%
-% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
-% Users can also go through the FAQs available on the journal's submission webpage.
-%
-% Steps to compile: latex, bibtex, latex latex
-%
-% For tracking purposes => this is v1.3 - March 2012
-
-\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax
+\documentclass{acm_proc_article-sp}
 
@@
 
-% Metadata Information
-% \acmVolume{9}
-% \acmNumber{4}
-% \acmArticle{39}
-% \acmYear{2010}
-% \acmMonth{3}
-
-% Copyright
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-%
-% % DOI
-% \doi{0000001.0000001}
-%
-% %ISSN
-% \issn{1234-56789}
-
 % Document starts
@@
 
-% Page heads
-\markboth{B. Berger et al.}{Computational biology in the 21st century: Algorithms that scale}
+\title{Computational biology in the 21st century: Algorithms that scale}
+\numberofauthors{3}
 
-% Title portion
-\title{Computational biology in the 21st century: Algorithms that scale}
-\author{BONNIE BERGER
-\affil{Massachusetts Institute of Technology}
-NOAH M. DANIELS
-\affil{Massachusetts Institute of Technology}
-Y. WILLIAM YU
-\affil{Massachusetts Institute of Technology}}
+\author{
+\alignauthor
+Bonnie Berger\\
+       \affaddr{Massachusetts Institute of Technology}\\
+       \affaddr{CSAIL and Department of Mathematics}\\
+       \affaddr{77 Massachusetts Ave}\\
+       \affaddr{Cambridge, MA 02139}\\
+       \email{bab@mit.edu}       
+\alignauthor
+Noah M. Daniels\\
+       \affaddr{Massachusetts Institute of Technology}\\
+       \affaddr{CSAIL and Department of Mathematics}\\
+       \affaddr{77 Massachusetts Ave}\\
+       \affaddr{Cambridge, MA 02139}\\
+       \email{ndaniels@mit.edu}
+\alignauthor
+Y. William Yu\\
+       \affaddr{Massachusetts Institute of Technology}\\
+       \affaddr{CSAIL and Department of Mathematics}\\
+       \affaddr{77 Massachusetts Ave}\\
+       \affaddr{Cambridge, MA 02139}\\
+       \email{ywy@mit.edu}
+}
 
+\maketitle
 \begin{abstract}
@@
 
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below. 
-%
-% \begin{CCSXML}
-% <ccs2012>
-%  <concept>
-%   <concept_id>10010520.10010553.10010562</concept_id>
-%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-%   <concept_significance>500</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10010520.10010575.10010755</concept_id>
-%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
-%   <concept_significance>300</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10010520.10010553.10010554</concept_id>
-%   <concept_desc>Computer systems organization~Robotics</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10003033.10003083.10003095</concept_id>
-%   <concept_desc>Networks~Network reliability</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-% </ccs2012>
-% \end{CCSXML}
-%
-% \ccsdesc[500]{Computer systems organization~Embedded systems}
-% \ccsdesc[300]{Computer systems organization~Redundancy}
-% \ccsdesc{Computer systems organization~Robotics}
-% \ccsdesc[100]{Networks~Network reliability}
-
-%
-% End generated code
-%
-
-% We no longer use \terms command
-%\terms{Design, Algorithms, Performance}
+% A category with the (minimum) three required fields
+\category{J.3}{Life and Medical Sciences}{Biology and genetics}
 
 \keywords{computational biology, database search, metagenomics, chemogenomics}
-
-\acmformat{Bonnie Berger, Noah M. Daniels,
-and Y. William Yu, 2015. Computational biology in the 21st century: Algorithms that scale.}
-% At a minimum you need to supply the author names, year and a title.
-% IMPORTANT:
-% Full first names whenever they are known, surname last, followed by a period.
-% In the case of two authors, 'and' is placed between them.
-% In the case of three or more authors, the serial comma is used, that is, all author names
-% except the last one but including the penultimate author's name are followed by a comma,
-% and then 'and' is placed before the final author's name.
-% If only first and middle initials are known, then each initial
-% is followed by a period and they are separated by a space.
-% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.
-
-\begin{bottomstuff}
-This work is supported by the National Institutes of Health, under
-grant GM108348.
-% This work is supported by the National Science Foundation, under
-% grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
-%
-% Author's addresses: G. Zhou, Computer Science Department,
-% College of William and Mary; Y. Wu  {and} J. A. Stankovic,
-% Computer Science Department, University of Virginia; T. Yan,
-% Eaton Innovation Center; T. He, Computer Science Department,
-% University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
-% (Current address) NASA Ames Research Center, Moffett Field, California 94035.
-\end{bottomstuff}
-
-\maketitle
 
@@
 \caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
-% YWY: My inclination is to write "(a) Moore's and (b) Kryder's laws". Possibly a matter of taste though, so do as you will.
 \label{fig:growth}
@@
 and analyze large libraries of data.
-% YWY: `omics' is the only single quote we use. Are we consistent?
 One approach to solving these challenges is to embrace cloud computing.
@@
 
-\section{Acknowledgments}
+\section{Acknowledgements}
+This work is supported by the National Institutes of Health, under
+grant GM108348.
 Y.W.Y. is supported by a Hertz Foundation Fellowship.
@@
 % Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
-% \bibliographystyle{abbrv}
+\bibliographystyle{abbrv}
 \bibliography{main}

### 2015-07-29 main.tex
Figure and table cleanup

@@
 
+\usepackage{array}
 \usepackage{caption}
@@
 % YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
-\begin{figure}[htb!]
+\begin{figure*}[htb!]
 \centering
@@
 \label{fig:growth}
-\end{figure}
+\end{figure*}
 
@@
 
-% YWY: Take spell-check lines out of figure labels! (bp, Samtools)
-\begin{figure}[htb!]
+\begin{figure*}[htb!]
 \centering
 \subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
-\includegraphics[width=2.6in]{assets/shotgun.png}
+\includegraphics[height=2.6in]{assets/shotgun.png}
 \label{fig:shotgun}
@@
 \subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
-\includegraphics[width=2.6in]{assets/snps.png}
+\includegraphics[height=2.6in]{assets/snps.png}
 \label{fig:snps}
@@
 \label{fig:ngs}
-\end{figure}
+\end{figure*}
 
@@
 Thus, not only do biological data exhibit redundancy, they also tend not to 
-inhabit anywhere near the entire feasible space (Figure~\ref{fig:clusters}).
+inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
 It seems that physical laws have constrained the data to a particular subspace
@@
 \centering
-\subfloat[Cartoon depiction of points in an arbitrary high-dimensional space, 
+\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
+\caption{Cartoon depiction of points in an arbitrary high-dimensional space, 
 as might arise from genomes generated by mutation and selection during the 
@@
 similarity search. Clusters cover the data points but do not cover unoccupied 
-regions of space.]{
-\includegraphics[width=2.8in]{assets/treepoints.pdf}
-\label{fig:clusters}
-}
-\subfloat[A query (green triangle) and two concentric search radii (red 
+regions of space.
+The green triangle represents a query, with two concentric search radii (red 
 circles) around it. Thanks to low fractal dimension, the large circle does
-not contain vastly more points than the small circle.]{
-\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
-\label{fig:fractal}
-}
-\caption[Data spaces.]{A cartoon illustration of a biological data space.}
+not contain vastly more points than the small circle.}
 \label{fig:dataspace}
@@
 entropy, which is relatively small compared to
-$|D|$, the number of entries in the database (Figure~\ref{fig:clusters}).
+$|D|$, the number of entries in the database (Figure~\ref{fig:dataspace}).
 In contrast, if the points were uniformly distributed about the Cartesian space,
@@
 $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
-number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:fractal}).
+number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:dataspace}).
 
@@
 proportional to
-\[
+\begin{displaymath}
     O\Bigg(
@@
      \Bigg) .
-\]
+\end{displaymath}
 
@@
 
+\newcolumntype{H}[1] {%
+>{\raggedright}%
+p{#1}}
+
 \begin{table}
@@
 NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
-\begin{tabular}{ccc}
+\tabcolsep=0.11cm
+\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
 \hline
@@
 \hline
-Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
+Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
 \hline
-Protein sequences (NCBI NR) & 5:1 & 1.6\\
+Protein~sequences (NCBI NR) & 5:1 & 1.6\\
 \hline
-Protein structure (PDB) & 10:1 & 2.5\\
+Protein~structure (PDB) & 10:1 & 2.5\\
 \hline
-Chemical structure (PubChem) & 11:1 & 0.2\\
+Chemical~structure (PubChem) & 11:1 & 0.2\\
 \hline

### 2015-07-29 main.tex
supplement we mightn't need.

@@
 As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
-(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types.
-Currently each sample requires seven hours to download. 
-% YWY: 7 hours for whom? Perhaps "even on an institutional connection"?
+(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer 
+researchers from about 80 institutions in a coordinated manner with the goal of 
+mapping the entire mutational landscape of 37 common cancer types. 
+Currently each sample requires seven hours to download even on an institutional 
+connection. 
 Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
@@
 \subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
-\includegraphics[height=2.6in]{assets/shotgun.png}
+\includegraphics[height=2.0in]{assets/shotgun.png}
 \label{fig:shotgun}
@@
 \subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
-\includegraphics[height=2.6in]{assets/snps.png}
+\includegraphics[height=2.0in]{assets/snps.png}
 \label{fig:snps}
@@
 This allows relationships between genes and diseases to be uncovered.]{
-\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
+\includegraphics[width=5.0in]{assets/ngs-pipeline.pdf}
 \label{fig:ngs-pipeline}
@@
 search for them in a protein database.
-% YWY: 16S isn't well introduced
 The difficulty is that the datasets required to shine any light on these 
@@
 upon here.
-
-Over the course of evolutionary time, structure is known to be more highly 
-conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
-does not change as rapidly.
-When the structure of a protein is known but its biological function or 
-evolutionary relationships are not, researchers may search for structurally
-similar proteins that are better studied~\cite{gibrat1996surprising}.
-Classical tools for this involve performing pairwise structural alignments to
-look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
-along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
-Matt~\cite{menke2008matt}.
-Due to the complexity of protein structures, these programs generally take 
-significant amounts of time,
-especially to align multiple structures.
-For example, the DALI webserver can take as much as an hour to return results
-for a single query.
-FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
-approximating structural alignments
-by instead comparing the `bag-of-words' from each structure.
-Analogous to a term-frequency vector in information retrieval, this bag-of-words
-indicates the abundance of particular, short structural motifs within a protein.
 
@@
 
-% In the realm of protein structure prediction, it has been
-% observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous
-% range of
-% possible dihedral angles, protein structures inhabit a smaller subspace, defined
-% by a narrower range of angles, referred to as
-% rotamers~\cite{dunbrack1993backbone}.
-% The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation
-% for the
-% purpose of sidechain packing (finding the lowest-energy conformation for the
-% sidechain of each amino acid), using tree width to reduce the search space and
-% simplify dependencies.
-
 \section{Conclusions and future prospects}

### 2015-07-29 supplement.tex
supplement we mightn't need.

\documentclass[article]

\begin{document}

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the `bag-of-words' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2015-07-29 main.tex
more cutting, and fix quotes

@@
 
-As mentioned earlier, function follows form, so in addition to
-sequence and expression, structure plays an important role in biological data science.
-Protein structure data, primarily determined through X-ray crystallography or
-nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
-each atom, in Angstroms, of a protein in three-dimensional space.
-This information is important to understanding how a protein functions, because
-proteins interact physically with other molecules and are geometrically constrained.
-Protein structure data are commonly stored in the Protein Data Bank (PDB); 
-each PDB entry has an associated resolution indicating the accuracy of the
-structural description.
-Thus a PDB entry at its most basic level contains a list of atoms, along with
-which amino acid they belong to and their spatial coordinates as floating-point
-numbers.
-
-Chemical compounds represent another source of structural data.
-Although generically much smaller than proteins, their structures are more general
-as they are not constrained in composition to amino acids.
+Chemical compounds represent a source of structural data.
 Determined by X-ray crystallography, NMR, electron microscopy and other 
@@
 the discovery of new 
-antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
+antibiotics~\cite{forsberg2012shared}), or it could be 
 the human gut, whose microbiome has been linked to human-health concerns 
 including Autism Spectrum Disorder~\cite{macfabe2012short}, 
-Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
+Crohn's Disease~\cite{gevers2014treatment}, and 
 obesity~\cite{greenblum2012metagenomic}.
@@
 of gene expression 
-data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
+data~\cite{raychaudhuri2000principal,alter2000singular}.
 Such data and its challenges have been the focus of other 
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an `approximate seed' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
 the space of possible matches, likewise pruning the search space; however, the
@@
 SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
-brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
+brings ideas from compressed sensing~\cite{candes2005decoding} 
 to gene expression analysis.
@@
 of July 
-2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
+2015, the NCBI's exhaustive `NR' database lists $6.9 \times 10^{7}$ sequences.
 This surprisingly lower number is not simply due to undersampling; thanks to 
@@
 One key insight related to redundancy
-%as to why the compressive omics approach allows computation 
-%that scales sublinearly in the size of the data 
 is that such data sets exhibit
@@
 
-Cluster-based search, as exemplified by ``compressive omics,'' can perform
+Cluster-based search, as exemplified by `compressive omics,' can perform
 approximate search within a radius $r$ of a query $q$ on a database $D$ with 
@@
 not necessarily, a power of two) used as `seeds' for longer sequence matches.
-CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
-a small Hamming or Levenshtein distance.
+CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
+one to its neighbors within a small Hamming or Levenshtein distance.
 The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

### 2015-07-29 supplement.tex
more cutting, and fix quotes

@@
 \begin{document}
+
+As mentioned earlier, function follows form, so in addition to
+sequence and expression, structure plays an important role in biological data science.
+Protein structure data, primarily determined through X-ray crystallography or
+nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
+each atom, in Angstroms, of a protein in three-dimensional space.
+This information is important to understanding how a protein functions, because
+proteins interact physically with other molecules and are geometrically constrained.
+Protein structure data are commonly stored in the Protein Data Bank (PDB); 
+each PDB entry has an associated resolution indicating the accuracy of the
+structural description.
+Thus a PDB entry at its most basic level contains a list of atoms, along with
+which amino acid they belong to and their spatial coordinates as floating-point
+numbers.
 

### 2015-07-31 main.tex
final tweaks for submission

@@
 
-\title{Computational biology in the 21st century: Algorithms that scale}
+\title{Computational biology in the 21st century:\\Algorithms that scale}
 \numberofauthors{3}
@@
 Amazon Web Services are also commonly used for computational biology 
-research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
+research and enterprise (e.g., DNAnexus)~\cite{schatz2010cloud}.
 However, while cloud computing frees researchers from maintaining their own
@@
 structure~\cite{berger2013computational} to speed up computation and save storage.
-This review focuses on cutting-edge algorithmic advances for dealing with the growth in 
+This Review focuses on cutting-edge algorithmic advances for dealing with the growth in 
 biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
@@
 abundance they occur (gene expression), and how they interact
-with both each other (interaction networks) and small chemical compound agonists.
+with both each other (interaction networks) and small chemical compound 
+agonists.
 
-% YWY: "Most" is an extremely strong claim. I believe you, but I could see people
-% arguing (e.g. all of electronic health records could be considered "biological")
-Most but not all biological data take the form of \emph{sequence} data, either 
+Much but not all biological data take the form of \emph{sequence} data, either 
 nucleotide sequences (using a four-letter alphabet representing the four 
@@
 
-While sequence is important, scientists are also interested in measuring gene expression, or the quantity of
-a particular gene product (usually a protein or more recently RNA) present in the cell.
-Gene expression data is useful for relating genotype to phenotype, and is 
-important in studying many diseases.
-Usually obtained through microarrays or RNA-seq technologies, expression data is
-quantitative, as each gene from a sample is associated with a numeric expression
-level, and high-dimensional, as many thousands of genes may be analyzed at a
-time.
-Resources such as NCBI’s Gene Expression Omnibus (GEO), which pool 
-together many diverse expression studies, now allow researchers to analyze 
-thousands of samples generated by others to obtain meaningful biological 
-insights. 
-More importantly, many of these biological insights can only be gleaned when 
-viewed in the context of hundreds or thousands of gene expression samples.
-
+As mentioned earlier, function follows form, so in addition to
+sequence and expression, structure plays an important role in biological data science.
 Chemical compounds represent a source of structural data.
@@
 possible because reads overlap and because the sequence is covered by many such
-overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
+overlapping reads~\cite{berger2013computational}; the \emph{de Bruijn graph} is a 
 commonly-used data structure.
@@
 complete genomes are catalogued.
-Likewise, WormBase~\cite{harris2014wormbase} and 
-FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
+Likewise, WormBase and 
+FlyBase are cataloguing many different species and
 strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
@@
 including Autism Spectrum Disorder~\cite{macfabe2012short}, 
-Crohn's Disease~\cite{gevers2014treatment}, and 
-obesity~\cite{greenblum2012metagenomic}.
+Crohn's Disease, and 
+obesity~\cite{kostic2014microbiome}.
 Metagenomics fundamentally asks what organisms are present, and, in the case
@@
 protein takes, and what biological function it performs.
-Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
+Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
 beyond the scope of this Review.
@@
 The homology search problem was previously believed to be solved;
-Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
+Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
 the standard tool for performing homology (similarity) search on databases of nucleotide and
@@
 
-While gene expression data lends itself to cluster analysis and probabilistic 
+While gene expression data (see Supplement) lends itself to cluster analysis 
+and probabilistic 
 approaches, the high dimensionality and noise in the data presents significant 
-challenges~\cite{nguyen2002tumor}.
+challenges~\cite{berger2013computational}.
 Principal Component Analysis has shown promise in reducing the dimensionality
 of gene expression 
-data~\cite{raychaudhuri2000principal,alter2000singular}.
+data~\cite{berger2013computational}.
 Such data and its challenges have been the focus of other 
@@
 Techniques for reference-based read mapping typically rely on algorithmic 
-approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
+approaches such as the Burrows-Wheeler transform (BWT)~\cite{berger2013computational}, 
 which provides efficient string 
 compression through a reversible transformation, while the FM-index data 
-structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
+structure~\cite{berger2013computational} is a compressed substring index, 
 based on the BWT, which provides efficient storage as well as fast search.
-BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
-Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
+BWA (Burrows-Wheeler Aligner)~\cite{berger2013computational} uses the BWT, while the 
+Bowtie2~\cite{berger2013computational} mapper further relies on the FM-index for 
 efficient mapping of NGS reads.
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{berger2013computational} use an `approximate seed' approach to indexing
 the space of possible matches, likewise pruning the search space; however, the
@@
 Even with these approaches, read mapping remains a significant bottleneck in
-genomic research~\cite{hach2010mrsfast}.
+genomic research~\cite{berger2013computational}.
 
@@
 
-Recent work on gene expession has explored additional ways to exploit the high-dimensional 
+Recent work on gene expression has explored additional ways to exploit the high-dimensional 
 structure of the data.
@@
 to gene expression analysis.
-Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
-are unlikely to be close in a lower-dimensional linear subspace.
-Thus, SPARCLE approximates a gene's expression profile as a linear combination
-of a few other genes' profiles, and then applies supervised learning to
-predict relationships between genes.
 Another recent and novel approach to exploiting the structure of gene expression
@@
 that polytope from the features most highly enriched at those vertices.
-In human breast tumors, and in mouse tissues, the expression data were
-well-described by tetrahedra whose vertices were enriched for different tumor
-types and biological functions, from which four distinct tasks are 
-inferred~\cite{hart2015inferring}.
 
@@
 alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
-IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank} 
+IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
 but performs a random walk on the product graph of two networks;
 IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
-PageRank-Nibble algorithm~\cite{andersen2006local}.
+PageRank-Nibble algorithm~\cite{berger2013computational}.
+NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
+conserved paths and conserved interaction clusters.
 Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
@@
 In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
-MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
+MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
 schemes aim to accelerate only the seed-to-reference matching step.
@@
 generalized and adapted to non-sequence spaces as well, and coined
-`compressive omics'~\cite{yu2015entropy}.
+`compressive omics.'
 One such example is chemogenomics.
@@
 Several approaches to compressing quality scores have been 
-presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality,yu2014traversing}.
+presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
 Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
@@
 This approach allows a general-purpose compressor to achieve tight compression.
-SCALCE~\cite{hach2012scalce} also relies on a `boosting' scheme, reordering
+SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
+reordering
 reads in such a way that a general-purpose compressor achieves improved
@@
 grant GM108348.
-Y.W.Y. is supported by a Hertz Foundation Fellowship.
+Y.W.Y. is supported by a Hertz Fellowship.
 

### 2015-07-31 supplement.tex
final tweaks for submission

@@
-\documentclass[article]
+\documentclass[12pt]{article}
+
+\title{Computational biology in the 21st century:\\Algorithms that scale\\Supplement}
+\date{}
 
@@
 
-As mentioned earlier, function follows form, so in addition to
-sequence and expression, structure plays an important role in biological data science.
+\maketitle
+
+\section{Gene expression}
+
+While sequence is important, scientists are also interested in measuring gene expression, or the quantity of
+a particular gene product (usually a protein or more recently RNA) present in the cell.
+Gene expression data is useful for relating genotype to phenotype, and is 
+important in studying many diseases.
+Usually obtained through microarrays or RNA-seq technologies, expression data is
+quantitative, as each gene from a sample is associated with a numeric expression
+level, and high-dimensional, as many thousands of genes may be analyzed at a
+time.
+Resources such as NCBI’s Gene Expression Omnibus (GEO), which pool 
+together many diverse expression studies, now allow researchers to analyze 
+thousands of samples generated by others to obtain meaningful biological 
+insights. 
+More importantly, many of these biological insights can only be gleaned when 
+viewed in the context of hundreds or thousands of gene expression samples~\cite{berger2013computational}.
+
+\section{Protein structure}
+
 Protein structure data, primarily determined through X-ray crystallography or

### 2015-09-10 main.tex
Some changes, and reviewer comments included

@@
 
-\title{Computational biology in the 21st century:\\Algorithms that scale}
+\title{Computational biology in the 21st century:\\The growth of compressive 
+omics}
 \numberofauthors{3}
@@
 
-% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
 \begin{figure*}[htb!]
@@
 features~\cite{berger2013computational}.
-As another example, clever indexing techniques such as the Burroughs-Wheeler 
+As another example, clever indexing techniques such as the Burrows-Wheeler 
 Transform (BWT) take advantage of aspects of sequence 
@@
 focus.
+
+\fbox{\parbox[t]{\linewidth}{
+
+(16S, shaped seeds, blocked
+reactions, chemogenomics, de Bruijn)
+\textbf{Definitions}
+
+\textbf{chemogenomics} This means this and that
+
+\textbf{metagenomics} This means another thing and so on and so forth blah blah blah blah blah.
+
+\textbf{metabolomics} And more.
+
+\textbf{shotgun sequencing} And more.
+
+\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database.
+
+\textbf{transcriptome} And more.
+
+\textbf{FASTQ} A common plain-text format for biological sequence data.
+
+\textbf{BLAST} And more.
+
+\textbf{16S}
+
+\textbf{de Bruijn graph}
+
+}}
 
@@
 
+Given a sequenced genome, the next natural questions ask what genes (genomic
+regions that code for proteins) are present, what structure each resulting
+protein takes, and what biological function it performs.
+Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
+beyond the scope of this Review.
+However, determining evolutionary relationships, structure, and function is
+at the heart of current research in computational biology.
+Since some organisms (known as \emph{model organisms}) are better studied than
+others, and evolution is known to conserve sequence, structure, and function,
+a powerful approach to determine these attributes is to search for similar
+sequences about which more is known. 
+This so-called \emph{homology search} entails searching for \emph{approximate}
+matches in databases of known gene or protein sequences.
+The homology search problem was previously believed to be solved;
+Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
+the standard tool for performing homology (similarity) search on databases of nucleotide and
+protein sequences.
+BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
+that might lead to longer matches, and greedily extends them, ultimately 
+producing a \emph{sequence alignment} between a query and each potential 
+database hit.
+However, BLAST's running-time scales linearly with the size of the database being
+searched, which is problematic as sequence databases continue to grow at a
+faster rate than Moore's law.
+
 On a potentially even larger scale is the growth of metagenomic data.
@@
 genera~\cite{janda200716s}.
-
-Given a sequenced genome, the next natural questions ask what genes (genomic
-regions that code for proteins) are present, what structure each resulting
-protein takes, and what biological function it performs.
-Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
-beyond the scope of this Review.
-However, determining evolutionary relationships, structure, and function is
-at the heart of current research in computational biology.
-Since some organisms (known as \emph{model organisms}) are better studied than
-others, and evolution is known to conserve sequence, structure, and function,
-a powerful approach to determine these attributes is to search for similar
-sequences about which more is known. 
-This so-called \emph{homology search} entails searching for \emph{approximate}
-matches in databases of known gene or protein sequences.
-The homology search problem was previously believed to be solved;
-Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
-the standard tool for performing homology (similarity) search on databases of nucleotide and
-protein sequences.
-BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
-that might lead to longer matches, and greedily extends them, ultimately 
-producing a \emph{sequence alignment} between a query and each potential 
-database hit.
-However, BLAST's running-time scales linearly with the size of the database being
-searched, which is problematic as sequence databases continue to grow at a
-faster rate than Moore's law.
 
@@
 DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
-`shaped seeds' instead of simple $k$-mer seeds to index the database.
+`shaped seeds'---essentially, $k$-mers with wildcards at specific positions---instead of simple $k$-mer seeds to index the database.
 DIAMOND demonstrates search performance three to four orders of magnitude faster
@@
 Mongoose~\cite{chindelevitch2014exact} is able to 
-identify and unblock many blocked reactions in current models without loss of 
-accuracy.
+identify and unblock many blocked reactions--those whose metabolites are out of 
+balance---in current models without loss of accuracy.
 
@@
 stands out.
-Any high-dimensional data set, such as amino acid sequences of a bounded 
-length, can be described as inhabiting a vector space.
-For example, given an alphabet of 20 amino acids and a length bounded at 
-500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
-in length, proteins are typically described as comprising \emph{domains} that 
-evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
-could
-enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
-of July 
-2015, the NCBI's exhaustive `NR' database lists $6.9 \times 10^{7}$ sequences.
-This surprisingly lower number is not simply due to undersampling; thanks to 
-evolution, only those protein
-sequences which exhibit useful biological function should survive, and most
+Far fewer biological sequences exist than could be enumerated, but even more so, those that exist tend to be highly similar to many others.
+Thanks to 
+evolution, only those genes
+which exhibit useful biological function survive, and most
 random sequences of amino acids would not be expected to form stable structures.
+Since two human genomes differ on average by only 0.1\%, a collection of 1,000 human genomes contain less than twice the unique information of a single genome~\cite{loh2012compressive}.
 Thus, not only do biological data exhibit redundancy, they also tend not to 
 inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
-It seems that physical laws have constrained the data to a particular subspace
-of the enumerable Cartesian space.
+It seems that physical laws---in this case, evolution---have constrained the data to a particular subspace of the enumerable Cartesian space.
 
@@
 
-Cluster-based search, as exemplified by `compressive omics,' can perform
+Cluster-based search, as exemplified by `compressive omics'---the use of compression to accelerate analysis---can perform
 approximate search within a radius $r$ of a query $q$ on a database $D$ with 
@@
     \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
-     \Bigg) .
+     \Bigg) ,
 \end{displaymath}
+where $B_D(q,r)$ refers to the set of points in $D$ contained within a ball of
+radius $r$ about a point $q$.
 
@@
 Biological data sets exhibit redundancy, and are constrained to subspaces by
-physical laws.
+physical laws; that is, the vast majority of enumerable sequences and 
+structures do not exist because they are not advantageous (or at least, have 
+not been selected for by evolution).
 This combination results in low fractal dimension and low metric entropy 
@@
 fractal dimension at typical search radii for four biological data sets.
+Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
 NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
@@
 (referred to as a homology table).
-CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
+CORA, like caBLAST~\cite{loh2012compressive} and 
+caBLASTP~\cite{daniels2013comprssive}, accelerates existing tools (in this 
+case, read
 mappers including BWA or Bowtie2) by allowing them to operate in a compressed
@@
 running time by an average of 10x.
+
+The compressive omics approach can, in some cases, come at the cost of accuracy.
+However, these cases are well-defined.
+Compressive omics never results in false positives (with respect to the na\"ive 
+search technique being accelerated), because the fine search phase applies the
+same comparison to the candidates as the na\"ive approach.
+Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
+never occur.
+However, in practice, non-metric distance functions are used, such as E-values
+in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
+In practice, these error rates are low, and recall better than 90\% has been
+demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}
 
@@
 
+\textbf{rewrite paragraph}
 While this Review has focused on the importance of algorithms that leverage the

### 2015-09-14 main.tex
Some types of data changes

@@
 machines.
-Thus, researchers are interested in DNA, RNA, and protein sequences, as well
-as the structures that RNA and proteins form.
-Researchers also study the resulting functions of RNA and proteins, in what
-abundance they occur (gene expression), and how they interact
-with both each other (interaction networks) and small chemical compound 
-agonists.
+Methods have been developed to gather biological data from every level of this process,
+including sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
+It is the success of these methods that has led to the massive influx of biological data we address today.
 
-Much but not all biological data take the form of \emph{sequence} data, either 
+\emph{Sequence} data, either 
 nucleotide sequences (using a four-letter alphabet representing the four
 DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
-representing the twenty standard amino acids).
-Sequence data are obtained in several ways, including mass spectrometry (which
-can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
-allowing scientists to also infer the expression of the gene to which it might translate).
-However, the greatest volume of sequence data available are genomic in nature and 
-come from next-generation sequencing (NGS) technologies.
+representing the twenty standard amino acids)
+are obtained in several ways.
+For protein and RNA sequence data, mass spectrometry---which
+can determine protein sequence, as well as interactions---and RNA-seq---which can determine RNA sequence and abundance, 
+allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
+However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
+For this Review, it is important to understand the structure of NGS sequence data, so we give a bit more description.
 
@@
 preparation.
-
-Importantly to this Review, Illumina sequencing and other NGS approaches
+Illumina sequencing and other NGS approaches
 such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
@@
 \emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
-These reads are read concurrently and must be reassembled in the correct order 
+These reads must be put together in the correct order 
 to piece together an entire genome.
@@
 available with some technologies (e.g., PacBio).
-This so-called `shotgun sequencing' is analogous to the problem of 
-reconstructing a book with all its pages torn out.
-However, these NGS reads overlap, and they are many in number, so 
-reconstructing an entire genome is feasible.
-
-Additionally, no sequencing technology is completely infallible, so sequencing machines also
+Because no sequencing technology is completely infallible, sequencing machines also
 provide a \emph{quality score} (or measure of the confidence in the DNA base
@@
 characters that encode the quality of the base call.
-A sequencing run will produce many of these reads, typically stored in a FASTQ
+A sequencing run will produce many overlapping reads, typically stored in a FASTQ
 file.
@@
 sequence and expression, structure plays an important role in biological data science.
-Chemical compounds represent a source of structural data.
+However, we are not interested in only RNA and protein structures;
+small chemical compounds represent an additional source of relevant structural data,
+as they often interact with their larger RNA and protein brethren.
 Determined by X-ray crystallography, NMR, electron microscopy and other 
@@
 
+%The assembly problem in so-called `shotgun sequencing' is analogous to the problem of 
+%reconstructing a book with all its pages torn out.
+%However, these NGS reads overlap, and they are many in number, so 
+%reconstructing an entire genome is feasible.
 Given DNA or RNA reads from NGS technologies, the first task is to 
-\emph{assemble} those fragments of sequence (which have associated quality 
-scores) into contiguous sequences.
-The task of \emph{de novo} assembly is beyond the scope of this Review, but is
-possible because reads overlap and because the sequence is covered by many such
-overlapping reads~\cite{berger2013computational}; the \emph{de Bruijn graph} is a 
-commonly-used data structure.
+\emph{assemble} those fragments of sequence into contiguous sequences.
+\emph{de novo} assembly is beyond the scope of this Review, but is
+possible because the sequence is covered by many
+overlapping reads~\cite{berger2013computational}; for this task, the \emph{de Bruijn graph}
+data structure is commonly used.
 Often, however, a \emph{reference genome} (or in the case of RNA, 

### 2015-09-15 main.tex
Reworked types of biological data (again)

@@
 Methods have been developed to gather biological data from every level of this process,
-including sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
-It is the success of these methods that has led to the massive influx of biological data we address today.
+resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins,
+Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on several examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.
 
@@
 However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
-For this Review, it is important to understand the structure of NGS sequence data, so we give a bit more description.
+To better understand the structure of NGS sequence data, we expand a little on NGS methodologies.
 
@@
 molecule end-to-end as one could read through a bound book.
-Instead, DNA molecules are chopped into many small fragments;
+Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
 from these fragments we generate 
@@
 as they often interact with their larger RNA and protein brethren.
-Determined by X-ray crystallography, NMR, electron microscopy and other 
-techniques, chemical compounds---many of which are useful pharmaceuticals---can be
-represented as labeled graphs (or equivalently, adjacency matrices).
-In such graphs, the vertices are atoms (labeled with a chemical element) and
-the edges are bonds between atoms (labeled with a bond type, such as single or
-double).
-Chemical compounds are stored in databases such as PubChem, and are often searched
-for by medicinal chemists as potential agonists for protein targets.
+Physical structures of molecules can be determined by X-ray crystallography, NMR, electron microscopy and other 
+techniques.
+Once determined, there are a variety of ways of representing these structures, from labeled graphs of molecular bonds to summaries of protein domains.
+These representations can then be stored in databases such as PubChem or the Protein Data Bank,
+and are often searched through for, e.g. potential small molecule agonists for protein targets.
+Importantly, as we will expand upon later in this Review, interesting biomolecules tend to be sparse and non-randomly distributed in many representational spaces, which can be used for accelerating the aforementioned searches.
 
@@
 These relationships and interactions play crucial roles in a systems-level understanding of biology.
+While standard network science tools have been employed in these analyses, metabolic pathways have additional flow constraints on their links;
+as with our other examples of interesting biological data types, these additional constraints on metabolic pathways allows for more targeted approaches to gain advantages over more general techniques.
 

### 2015-09-15 main.tex
Minor changes and moved paragraph on gene expression to types of biological data

@@
 
+While measuring abundance to generate gene expression data (see Supplement) lends itself to cluster analysis 
+and probabilistic 
+approaches, the high dimensionality and noise in the data presents significant 
+challenges~\cite{berger2013computational}.
+Principal Component Analysis has shown promise in reducing the dimensionality
+of gene expression 
+data~\cite{berger2013computational}.
+Such data and its challenges have been the focus of other 
+articles~\cite{berger2013computational}, and thus will be only lightly touched 
+upon here.
+
 As mentioned earlier, function follows form, so in addition to
@@
 
-%The assembly problem in so-called `shotgun sequencing' is analogous to the problem of 
-%reconstructing a book with all its pages torn out.
-%However, these NGS reads overlap, and they are many in number, so 
-%reconstructing an entire genome is feasible.
 Given DNA or RNA reads from NGS technologies, the first task is to 
 \emph{assemble} those fragments of sequence into contiguous sequences.
+The assembly problem is analogous to the problem of 
+reconstructing a book with all its pages torn out.
 \emph{de novo} assembly is beyond the scope of this Review, but is
@@
 this reference (Figure~\ref{fig:ngs-pipeline}).
+Continuing the book analogy, it's much easier to reconstruct a book with all its pages torn out when one has another (perhaps imperfect) copy of that book to match pages to.
 Mapping allows the differences between the newly-sequenced genome and the 
@@
 genera~\cite{janda200716s}.
-
-While gene expression data (see Supplement) lends itself to cluster analysis 
-and probabilistic 
-approaches, the high dimensionality and noise in the data presents significant 
-challenges~\cite{berger2013computational}.
-Principal Component Analysis has shown promise in reducing the dimensionality
-of gene expression 
-data~\cite{berger2013computational}.
-Such data and its challenges have been the focus of other 
-articles~\cite{berger2013computational}, and thus will be only lightly touched 
-upon here.
 

### 2015-09-19 main.tex
Subtitle change per Bonnie's suggestion
Minor changes to types of biological data

@@
 
-\title{Computational biology in the 21st century:\\The growth of compressive 
-omics}
+\title{Computational biology in the 21st century:\\Scaling with compressive algorithms}
 \numberofauthors{3}
@@
 
-\section{Types and sources of biological data}
+\section{Types of biological data}
 
@@
 
-\emph{Sequence} data, either 
+Sequence data, either 
 nucleotide sequences (using a four-letter alphabet representing the four

### 2015-09-20 main.tex
Very compressed single network science section

@@
 When examining more complex phenomena than single proteins or compounds, we
-often look to synthesize things together.
+often look to synthesize things together into a systems-level understanding of biology.
 To that end, we frequently use
@@
 pathways~\cite{berger2013computational}.
-These relationships and interactions play crucial roles in a systems-level understanding of biology.
-While standard network science tools have been employed in these analyses, metabolic pathways have additional flow constraints on their links;
-as with our other examples of interesting biological data types, these additional constraints on metabolic pathways allows for more targeted approaches to gain advantages over more general techniques.
+While standard network science tools have been employed in these analyses---
+e.g. several approaches make use of diffusion/random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
+they are often paired with more specific biological data,
+as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
+Other tools solve more specifically biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
+aims to identify and unblock metabolic networks.
+However, given its breadth, biological network science is beyond the scope of this Review, so we will not focus on that here.
 
+%NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
+%conserved paths and conserved interaction clusters.
+%IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
+%PageRank-Nibble algorithm~\cite{berger2013computational}.
+%Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
+%which defines a distance metric between two nodes based on random walks.
+%DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
+%functions when applied to standard approaches such as neighbor voting for function annotation.
+%Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
+%analogue of principle component analysis; DCA reduces the dimensionality of a
+%network in such a way that function annotation can be performed in the reduced 
+%space, significantly decreasing running time while still providing competitive
+%prediction accuracy.
 \section{Challenges with biological data}
@@
 
-With biological networks, challenges lie in integrating such data in order to 
-deduce gene function, as well as uncover disease and metabolic pathways.
-Several approaches have taken advantage of the particular topologies of 
-biological
-networks, such as diffusion-based approaches, which explore the topology of
-networks through random walks.
-In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
-interact with
-many partners; these interactions do not imply shared function as strongly as
-non-hub interactions; such information needs to be teased apart.
-As new functional annotations and interactions are discovered,
-the number and size of these networks grows, making related problems more 
-intractable~\cite{berger2013computational}.
-
 The continued ability to store, search, and analyze these growing data sets hinges on
@@
 cannnot be searched on a laptop computer with current tools such as SMSD.
-
-An area of great interest in integrating biological network data is in global multiple network 
-alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
-IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
-but performs a random walk on the product graph of two networks;
-IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
-PageRank-Nibble algorithm~\cite{berger2013computational}.
-NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
-conserved paths and conserved interaction clusters.
-Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
-which defines a distance metric between two nodes based on random walks.
-DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
-functions when applied to standard approaches such as neighbor voting for function annotation.
-Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
-analogue of principle component analysis; DCA reduces the dimensionality of a
-network in such a way that function annotation can be performed in the reduced 
-space, significantly decreasing running time while still providing competitive
-prediction accuracy.
-Using a compressed representation of metabolic networks, 
-Mongoose~\cite{chindelevitch2014exact} is able to 
-identify and unblock many blocked reactions--those whose metabolites are out of 
-balance---in current models without loss of accuracy.
 

### 2015-09-20 main.tex
Moved Quartz section into CORA paragraph

@@
 The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
+As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
+can also be used for storage compression of NGS quality scores associated with base calls.
+Unlike many other approaches to compressing quality scores in the literature
+~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp},
+Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
+bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
+Because Quartz injects information from the distribution of $l$-mers in the target genome in this lossy compression, it demonstrates not only improved compression over competing
+approaches, but slightly improves the accuracy of downstream variant-calling.
 
@@
 \section{Storage compression}
-
-Along with a genome sequence, modern sequencing systems provide a measure of
-the quality of, or confidence in, the nucleotide base read at every position.
-While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
-bits per position, these quality scores occupy a larger range, na\"ively
-requiring six or seven bits per position.
-Thus, quality scores offer low-hanging fruit with respect to compression for 
-storage and transmission of sequence data.
-Several approaches to compressing quality scores have been 
-presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
-Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
-compression, by bounding the likelihood that a quality score is informative.
-Importantly, Quartz modifies the FASTQ file of the reads and quality scores, and thus can be
-readily incorporated into existing sequence analysis pipelines.
-Surprisingly, Quartz not only demonstrates improved compression over competing
-approaches, but slightly improves the accuracy of downstream variant-calling.
+\textbf{TODO(NMD): remove/re-integrate section}.
 

### 2015-09-20 main.tex
Reworded last paragraph to be a bit more upbeat

@@
 
-\textbf{rewrite paragraph}
-While this Review has focused on the importance of algorithms that leverage the
-structure of data to cope with the increased throughput in computational 
-biology, it is also essential that techniques continue to make better 
+The massive influx of biological data offers up the possibility of better
 predictions and models that demonstrate improved fidelity to wet-lab 
 experiments.
+It is through the development of algorithms that leverage the
+structure of the data that we aim to cope with this increased throughput in computational 
+biology and realize that goal.
 
@@
 grant GM108348.
-Y.W.Y. is supported by a Hertz Fellowship.
+Y.W.Y. is also supported by a Hertz Fellowship.
 

### 2015-09-20 main.tex
Fixed first sentence of section 5

@@
 
-Fortunately, biological data has unique structure \cite{yu2015entropy}, which we 
-will take advantage of to perform search that scales sublinearly in the size of the database~(Section~\ref{omics}).
+Fortunately, biological data has unique structure, which we 
+take advantage of to perform search that scales sublinearly in the size of the database~(Section~\ref{omics}) \cite{yu2015entropy}.
 The first critical observation is that

### 2015-09-20 main.tex
Minor edits, and storage compression moved.

@@
 While standard network science tools have been employed in these analyses---
-e.g. several approaches make use of diffusion/random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
+e.g. several approaches make use of diffusion or random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
 they are often paired with more specific biological data,
@@
 aims to identify and unblock metabolic networks.
-However, given its breadth, biological network science is beyond the scope of this Review, so we will not focus on that here.
+However, given its breadth, biological network science is beyond the scope of this Review, so we will not discuss it further.
 
@@
 only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
-Even with these approaches, read mapping remains a significant bottleneck in
+Even with these approaches, read mapping remains a significant bottlenecks in
 genomic research~\cite{berger2013computational}.
+
+Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
+Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
+similar reads (those that share a common 
+substring) together into `buckets', allowing that common substring to now be 
+removed and treated as the bucket label, so that each read in the compressed 
+representation comprises only its unique differences from the bucket label.
+This approach allows a general-purpose compressor to achieve tight compression.
+SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
+reordering
+reads in such a way that a general-purpose compressor achieves improved
+compression ratios.
 
@@
 
-\section{The age of compressive omics}\label{omics}
+\section{The age of compressive algorithms}\label{omics}
 
-We are entering the age of compressive omics, 
+We are entering the age of compressive algorithms, 
 which makes use of this completely different paradigm for the structure of biological data.
@@
 CORA, like caBLAST~\cite{loh2012compressive} and 
-caBLASTP~\cite{daniels2013comprssive}, accelerates existing tools (in this 
+caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
 case, read
@@
 
-\section{Storage compression}
-\textbf{TODO(NMD): remove/re-integrate section}.
-
-Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
-similar reads (those that share a common 
-substring) together into `buckets', allowing that common substring to now be 
-removed and treated as the bucket label, so that each read in the compressed 
-representation comprises only its unique differences from the bucket label.
-This approach allows a general-purpose compressor to achieve tight compression.
-SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
-reordering
-reads in such a way that a general-purpose compressor achieves improved
-compression ratios.
-
 \section{Conclusions and future prospects}

### 2015-09-21 main.tex
cite de bruijn

@@
 
-(16S, shaped seeds, blocked
-reactions, chemogenomics, de Bruijn)
 \textbf{Definitions}
@@
 \textbf{BLAST} And more.
-
-\textbf{16S}
-
-\textbf{de Bruijn graph}
 
@@
 overlapping reads~\cite{berger2013computational}; for this task, the \emph{de Bruijn graph}
-data structure is commonly used.
+data structure is commonly used~\cite{bruijn1946combinatorial}.
 Often, however, a \emph{reference genome} (or in the case of RNA, 
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} and mrsFAST \cite{berger2013computational} use an `approximate seed' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
 the space of possible matches, likewise pruning the search space; however, the

### 2015-09-21 main.tex
new conclusion last para

@@
 approaches, the high dimensionality and noise in the data presents significant 
-challenges~\cite{berger2013computational}.
+challenges.
 Principal Component Analysis has shown promise in reducing the dimensionality
 of gene expression 
-data~\cite{berger2013computational}.
+data.
 Such data and its challenges have been the focus of other 
@@
 as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
-Other tools solve more specifically biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
-aims to identify and unblock metabolic networks.
+Other tools solve other biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
+analyzes metabolic networks.
 However, given its breadth, biological network science is beyond the scope of this Review, so we will not discuss it further.
@@
 Techniques for reference-based read mapping typically rely on algorithmic 
-approaches such as the Burrows-Wheeler transform (BWT)~\cite{berger2013computational}, 
+approaches such as the Burrows-Wheeler transform (BWT), 
 which provides efficient string 
 compression through a reversible transformation, while the FM-index data 
-structure~\cite{berger2013computational} is a compressed substring index, 
-based on the BWT, which provides efficient storage as well as fast search.
-BWA (Burrows-Wheeler Aligner)~\cite{berger2013computational} uses the BWT, while the 
-Bowtie2~\cite{berger2013computational} mapper further relies on the FM-index for 
-efficient mapping of NGS reads.
+structure is a compressed substring index, 
+based on the BWT, which provides efficient storage as well as fast search~\cite{berger2013computational}.
+BWA (Burrows-Wheeler Aligner)uses the BWT, while the 
+Bowtie2 mapper further relies on the FM-index for 
+efficient mapping of NGS reads~\cite{berger2013computational} .
 The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
@@
 (referred to as a homology table).
-CORA, like caBLAST~\cite{loh2012compressive} and 
+CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
 caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
@@
 Unlike many other approaches to compressing quality scores in the literature
-~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp},
+~\cite{bonfield2013compression,ochoa2013qualcomp},
 Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
@@
 
-The massive influx of biological data offers up the possibility of better
-predictions and models that demonstrate improved fidelity to wet-lab 
-experiments.
-It is through the development of algorithms that leverage the
-structure of the data that we aim to cope with this increased throughput in computational 
-biology and realize that goal.
+Biological data science is unique for two primary reasons: biology itself---even molecular 
+biology---predates the information age, and 
+``nothing in biology makes sense except in light of evolution''~\cite{dobzhansky1973nothing}.
+Not only have biologists developed a diverse array of experimental techniques, but the data
+derive from astoundingly complex processes that themselves are driven by evolution.
+It is through the development of algorithms that leverage the structure of biological data that we can
+make sense of biology in light of evolution.
 

### 2015-09-21 main.tex
Hopefully ready to submit.

@@
 
-\textbf{chemogenomics} This means this and that
+\textbf{chemogenomics} Computational study of drugs and their targets based on chemical structure and function
 
-\textbf{metagenomics} This means another thing and so on and so forth blah blah blah blah blah.
+\textbf{metagenomics} Study of the many genomes that make up a particular environment
 
-\textbf{metabolomics} And more.
+\textbf{shotgun sequencing} Modern genomic sequencing, which chops DNA into many short pieces
 
-\textbf{shotgun sequencing} And more.
+\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database
 
-\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database.
+\textbf{transcriptome} Transcribed RNA from a genome, which results in protein production
 
-\textbf{transcriptome} And more.
-
-\textbf{FASTQ} A common plain-text format for biological sequence data.
-
-\textbf{BLAST} And more.
+\textbf{BLAST} Standard biological sequence similarity search tool
 
@@
 This structure ultimately determines a protein's function within the cell.
-Certain kinds of RNA also function, sometimes with proteins, as cellular 
+Certain kinds of RNA also function as cellular 
 machines.
 Methods have been developed to gather biological data from every level of this process,
-resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins,
-Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on several examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.
+resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
+Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.
 
@@
 For protein and RNA sequence data, mass spectrometry---which
-can determine protein sequence, as well as interactions---and RNA-seq---which can determine RNA sequence and abundance, 
+can determine protein sequence and interactions---and RNA-seq---which can determine RNA sequence and abundance, 
 allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
 However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
-To better understand the structure of NGS sequence data, we expand a little on NGS methodologies.
+To better understand the structure of NGS sequence data, we will expand on NGS methodologies.
 
@@
 characters that encode the quality of the base call.
-A sequencing run will produce many overlapping reads, typically stored in a FASTQ
-file.
+A sequencing run will produce many overlapping reads.
 
@@
 Crohn's Disease, and 
-obesity~\cite{kostic2014microbiome}.
+obesity.
 Metagenomics fundamentally asks what organisms are present, and, in the case
@@
 the maximal common subgraph (MCS), an NP-hard problem.
-However, there are once again an increasing number of such chemical compounds
+However, there are an increasing number of such chemical compounds
 to search; the NCBI's PubChem database has grown from 31 million compounds in
@@
 reads in such a way that a general-purpose compressor achieves improved
-compression ratios.
+compression.
 
@@
 inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
-It seems that physical laws---in this case, evolution---have constrained the data to a particular subspace of the enumerable Cartesian space.
+It seems that physical laws---in this case, evolution---constrain the data to a particular subspace of the Cartesian space.
 
@@
 \centering
-\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
+\includegraphics[width=2.5in]{assets/treepoints-fractal.png}
 \caption{Cartoon depiction of points in an arbitrary high-dimensional space, 
@@
 \caption{Metric entropy ratio (ratio of clusters to entries in database) and
-fractal dimension at typical search radii for four biological data sets.
+fractal dimension at typical search radii for four data sets.
 Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
-NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
+NCBI's non-redundant `NR' protein and `NT' nucleotide sequence databases are from June, 2015. Protein Data Bank (PDB) is from July, 2015. PubChem is from October, 2013.}
 \tabcolsep=0.11cm
@@
 
-\section{The age of compressive algorithms}\label{omics}
+\section{The age of compressive\\ algorithms}\label{omics}
 
@@
 that represent unique data.
-Any representative sequence that is within some threshold of the query is then expanded 
-into all the similar sequences it represents; the fine search is then over this
+Any representative sequence within some threshold of the query is then expanded 
+into all similar sequences it represents; the fine search is over this
 (typically small) subset of the original database.
-This approach provides orders-of-magnitude run-time improvements to BLAST search on 
+This approach provides orders-of-magnitude run-time improvements to BLAST 
 nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
-data; moreover, these run-time improvements increase as databases grow.
+search; these run-time improvements increase as databases grow.
 
@@
 bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
-Because Quartz injects information from the distribution of $l$-mers in the target genome in this lossy compression, it demonstrates not only improved compression over competing
+Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
 approaches, but slightly improves the accuracy of downstream variant-calling.
@@
 In practice, these error rates are low, and recall better than 90\% has been
-demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}
+demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}.
 
-\section{Conclusions and future prospects}
+\section{Conclusions}
 
@@
 outside of biology are likely to bear fruit. 
-In turn, it is also likely that algorithmic ideas developed within computational biology
+It is also likely that algorithmic ideas developed within computational biology
 will become useful in other fields experiencing a data deluge, such as astronomy

### 2015-09-21 response.tex
response letter

\documentclass[12pt]{letter}
\setlength{\textwidth}{6.5 in}
\setlength{\textheight}{8.0in}
\setlength{\oddsidemargin}{0in}
\setlength{\topmargin}{-0.1in}
\addtolength{\textheight}{0.9in}
%\addtolength{\voffset}{-.5in}
%advance\textheight by 1.5in
%\advance\topmargin by 1.0in
\address{}
\signature{\hspace*{.1in} Bonnie Berger}

\begin{document}
\begin{letter}{}
\opening{Dear Dr. Shamir,}

\medskip

We thank you and the anonymous reviewer for insightful and thoughtful feedback
on our paper, which we have re-titled ``Computational biology in the 21st century: Scaling with compressive algorithms.'' We believe we have addressed
all major and minor points made by both you and and the reviewer.
Below, we address each point in turn.

\emph{(1) regarding the uneven background: A good solution is to include a box of basic biology and bioinformatics terms, rather than putting them inside the text. (But do keep in mind the total length constraint..)}

We have incorporated a box for definitions.

\emph{(2-3) I fully agree that the focus on compressive algorithms should be improved and that title should be more specific.}

We have changed the title, and attempted to narrow the focus to compressive algorithms.

\emph{a. Numerous references are made to [4], sometimes several times within the same paragraph. I realize this was a way to get around the limit on citations, but the result is odd. One solution would be to cite all the original references but split some of them to the supplement.}

We believe we have improved this situation, not by moving references to the supplement, but by pushing the reference to the end when needed, and by citing
the original paper when possible (we freed up some space for references by narrowing the focus of the paper).

\emph{b. page 3 right, the sentence starting ``A second way is to perform homology'' is vague and would be lost for the CACM audience.}

Homology search is now defined in the box, and we have improved the language here.

\emph{c. page 5 right, what does ``unblock many blocked reactions'' mean?}

As part of the re-focusing of the paper, we limit discussion of this and other network approaches to a brief mention, and this terminology is no longer present.

\emph{d. page 6 left ``compressive genomics'' is used here and is only defined later. B_D is not defined. The bottom paragraph was vague for me. }

We have defined B_D, rewritten the paragraph, and more properly introduced the term ``compressive genomics.''

\emph{e. page 6 Table 1 please explain the meaning of the results in the caption. }

We have expanded the caption.

\emph{f. page 7 left please discuss the issue of loss of solutions due to the compression method.}

We have added a paragraph discussing the accuracy limitations of compressive omics.

\emph{g. some jargon used and not defined: transcriptome, FASTQ, caBLAST, caBLASTP}

We have added definitions, either to the box (transcriptome and BLAST) or in the text (caBLAST/caBLASTP). In our opinion, the mention of FASTQ was a needless detail, so we have removed it entirely from the paper.

\emph{Review of Berger et al. "Computational biology in the 21st century: Algorithms
that scale"}

\emph{This manuscript reviews computational problems that are faced in biology now
that that discipline is awash in huge amounts of data.  The manuscript
generally focuses on problems in sequence analysis (mapping, assembly, gene
expression, metagenomics, etc.) but also touches on chemical structure
databases and problems biological networks. This topic is of crucial importance
to biology, computational biology, and --- I would argue --- computer science,
and so the article is very timely.  The authors are leading researchers in
developing fast algorithms for these problems and have produced many of the
techniques cited in this review.}

\emph{The statements made in the manuscript appear accurate to me, and as an expert
in computational methods for biology the manuscript was understandable.}

\emph{However, there are a number of stylistic weaknesses and weaknesses in choice of
material that could be fixed in a revised version:}

\emph{[1] It is not clear from the writing who the audience is for this article. On
the one hand common biological terms such as ``model organism'' are defined,
while more complex concepts and jargon are not (16S, shaped seeds, blocked
reactions, chemogenomics, de Bruijn). I suspect it would be difficult for a
general reader of CACM to follow some of this manuscript. Similarly, I think
the manuscript does not do enough to convince the general reader of the
interest or importance of the topics described.}

We have attempted to address this completely valid criticism in two ways: by
re-focusing the paper, some of these concepts (blocked reactions, 16S) do not come up. We have defined those terms that are still present. Moreover, we hope we have done a better job of emphasizing the importance of the topics at hand.

\emph{[2] Though it is admirable to try to cover the breadth of biological data
types, the flow of the article is often somewhat disjointed as the context is
switched from genomics, to networks, to chemical databases. }

We have narrowed the focus of the article; networks are only mentioned briefly in order to explain that they are outside the scope of the article, and we believe we have improved the connections between the disparate subjects (genomics, structure, and chemical databases).

\emph{[3] It isn't clear what the overall point that the authors are trying to make
is.  There are some clear sub-points (use the structure of the data, a
hierarchical algorithmic approach often works well), but the article lacks a
strong, novel point of view.  This starts from the subtitle: "Algorithms that
scale". This has been a goal of computer science since its start. The article
would benefit from the development of a more specific story, a more decisive
point of view, and a stronger backbone. As just an example, a title or subtitle
such as ``The success of compressive algorithms'' and the corresponding changes
to make that the early and complete focus of the manuscript would result in a
stronger, more interesting article.}

In addition to a new title, we have endeavored to strengthen the story regarding compressive algorithms, both by reorganizing sections of the paper and by narrowing the focus.

\emph{[4] More text could be spent synthesizing what is special and different about
biological ``big data'' than other (sometimes larger) data collections in other
disciplines.  Relatedly, the introduction spends a lot of time arguing for the
need for faster algorithms, which I think is a settled point nearly everywhere
in computer science. More interesting would be a more detailed discussion of
the situations in biology where faster algorithms are most crucial (this is
done with read mapping, but there must be others).}

Within the space limitations, we believe we have addressed this. We have rewritten the last paragraphs of the conclusion, in particular, to discuss how biological data is unique.

Minor points:

\emph{[5] If kept, the paragraph in section 3 starting with ``Given a sequenced
genome...'' should come before the paragraph that precedes it.}

We have made this change.

\emph{[6] Burroughs-Wheeler is misspelled in section 1. (It should be Burrows)}

We have corrected this error.

\emph{[7] The sample size argument of section 5, paragraph 2 [there are a possible
20^500 proteins but only 6.9 x 10^7 are observed] is not persuasive. Such
``search-space-size'' arguments are nearly always vapid.  This sentence is one of
a over
3984137914278306537107946300187788156651883090392267368064424070371960737746809881430938446547647791637956210590388569173298650466385810245792657795213961405107689148645376
possible strings of characters of this length. But I had to consider very few
of them to write it. I would find a more specific argument about evolution,
relatedness of species, or something else to argue that the low dimensionality
is surprising.}

We have attempted to produce a less vapid argument, by focusing on the redundancy inherent in (for example) multiple genomes.

\emph{[8] The last two paragraphs of the manuscript are somewhat disappointing. The
penultimate paragraph positions computational biology as a derivative field, an
applier of ideas from the broader CS community. The last paragraph deflates the
hope of computation driving science with the appeal to requiring wet-lab
experiments.  Neither paragraph is wrong, but their emphasis does not serve
computational biology well, in my opinion. }

We thank the reviewer for pointing this out; indeed, it was a rather dismal conclusion to the article. We have completely rewritten the conclusion after meditating on what distinguishes biological ``big data'' from other sources of big data.

\closing{\hspace*{.5in}Sincerely,}
\end{letter} \end{document}

### 2015-09-21 main.tex
remove 16S stuff

@@
 questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
-complex than the 16S metagenomic studies that merely classify the bacteria 
-based on 
-their divergence with respect to a specific gene (namely, the 16S ribosomal 
-subunit).
+complex than standard genomic datasets.
 The massive data results in major identification 
@@
 Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
-similar reads (those that share a common 
+similar reads (those that share a common short
 substring) together into `buckets', allowing that common substring to now be 

### 2015-09-22 main.tex
Bonnie's edits

@@
 
-\keywords{computational biology, database search, metagenomics, chemogenomics}
+\keywords{computational biology, database search, metagenomics, \\chemogenomics}
 
@@
 to generating biological data; next-generation sequencing, mass spectrometry,
-microarrays, nuclear magnetic resonance spectroscopy, and other high-throughput 
+microarrays, cryo-electron microscopy, and other high-throughput 
 approaches have led to an explosion of data.
@@
 power was still matching the growth rate of genomic data.
-However, the sequencing technology used for the Human Genome Project--Sanger
-sequencing--was supplanted around 2004, with the advent of what is now known as 
-next-generation sequencing.
+However, the sequencing technology used for the Human Genome Project---Sanger
+sequencing---was supplanted around 2004, with the advent of what is now known 
+as next-generation sequencing.
 The material costs to sequence a genome have plummeted in the past
@@
 While it is now possible to bring sequencing equipment and limited computing 
-resources to remote sites, internet connectivity is still highly constrained.
-Thus, accessing cloud resources for analytics may not be possible.
+resources to remote sites, internet connectivity is still highly constrained;
+accessing cloud resources for analytics may not be possible.
 
@@
 acids which singly or in complexes are known as proteins.
-Proteins fold into complex, low-energy structures, which function as
+Proteins fold into sophisticated, low-energy structures, which function as
 cellular machines; the DNA sequence determines the amino acid sequence,
@@
 are obtained in several ways.
-For protein and RNA sequence data, mass spectrometry---which
+For both protein and RNA sequence data, mass spectrometry---which
 can determine protein sequence and interactions---and RNA-seq---which can determine RNA sequence and abundance, 
@@
 method for reading a genome.
-Though slow and labor intensive, Sanger sequencing has the advantage of being
-well-tested and is still used to validate NGS results.
-More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
-beginning with Illumina's `sequencing by synthesis,' which allows vastly 
+More recently, however, NGS approaches,
+beginning with Illumina's `sequencing by synthesis,' have enabled vastly 
 greater throughput due to massive parallelism, low cost, and simple sample
@@
 as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
-Other tools solve other biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
+Other tools solve other biological problems, such as MONGOOSE~\cite{chindelevitch2014exact}, which 
 analyzes metabolic networks.
@@
 reconstructing a book with all its pages torn out.
-\emph{de novo} assembly is beyond the scope of this Review, but is
+\emph{De novo} assembly is beyond the scope of this Review, but is
 possible because the sequence is covered by many
@@
 this reference (Figure~\ref{fig:ngs-pipeline}).
-Continuing the book analogy, it's much easier to reconstruct a book with all its pages torn out when one has another (perhaps imperfect) copy of that book to match pages to.
+Continuing the book analogy, it is much easier to reconstruct a book with all its pages torn out when one has another (perhaps imperfect) copy of that book to match pages to.
 Mapping allows the differences between the newly-sequenced genome and the 
@@
 The homology search problem was previously believed to be solved;
-Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
+Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} has been
 the standard tool for performing homology (similarity) search on databases of nucleotide and
@@
 based on the BWT, which provides efficient storage as well as fast search~\cite{berger2013computational}.
-BWA (Burrows-Wheeler Aligner)uses the BWT, while the 
+BWA (Burrows-Wheeler Aligner) uses the BWT, while the 
 Bowtie2 mapper further relies on the FM-index for 
@@
 when mapping reads to a reference genome.
-Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
+Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to index
 the space of possible matches, likewise pruning the search space; however, the
@@
 only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
-Even with these approaches, read mapping remains a significant bottlenecks in
+Even with these approaches, read mapping remains a significant bottleneck in
 genomic research~\cite{berger2013computational}.
@@
 Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
-similar reads (those that share a common short
-substring) together into `buckets', allowing that common substring to now be 
+similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
 removed and treated as the bucket label, so that each read in the compressed 
 representation comprises only its unique differences from the bucket label.
-This approach allows a general-purpose compressor to achieve tight compression.
+This approach allows a general-purpose compressor to achieve better compression.
 SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
@@
 DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
-`shaped seeds'---essentially, $k$-mers with wildcards at specific positions---instead of simple $k$-mer seeds to index the database.
+`shaped seeds'---essentially, $k$-mers of length 15--24 with wildcards at 9--12 specific positions---instead of simple $k$-mer seeds to index the database.
 DIAMOND demonstrates search performance three to four orders of magnitude faster
-than BLASTX, but it is still linear in the size of the database being searched.
+than BLASTX, but still linear in the size of the database being searched.
 
@@
 \label{fractal}
-\caption{Metric entropy ratio (ratio of clusters to entries in database) and
+\caption{Metric-entropy ratio (ratio of clusters to entries in database) and
 fractal dimension at typical search radii for four data sets.
-Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
+Metric-entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
 NCBI's non-redundant `NR' protein and `NT' nucleotide sequence databases are from June, 2015. Protein Data Bank (PDB) is from July, 2015. PubChem is from October, 2013.}
@@
 \hline
-Data set & Metric entropy ratio & Fractal dimension \\        
+Data set & Metric-entropy ratio & Fractal dimension \\        
 \hline
@@
 We are entering the age of compressive algorithms, 
-which makes use of this completely different paradigm for the structure of biological data.
+which make use of this completely different paradigm for the structure of biological data.
 Seeking to take advantage of the redundancy inherent in genomic sequence data, 
@@
 performed in the compressed representation.
-Daniels et al. \cite{daniels2013compressive} further demonstrated that
-compressive acceleration takes advantage of increasing redundancy in the
-data over time.
-
 Compressive genomics is based on the concept of \emph{compressive acceleration},
@@
 mappers including BWA or Bowtie2) by allowing them to operate in a compressed
-space, and relying on a coarse and a fine phase.
+space, and relies on a coarse and a fine phase.
 In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
@@
 Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
-approaches, but slightly improves the accuracy of downstream variant-calling.
+approaches, but slightly improves the accuracy of downstream 
+variant-calling~\cite{yu2015quality}.
 
@@
 never occur.
-However, in practice, non-metric distance functions are used, such as E-values
+Yet, in practice, non-metric distance functions are used, such as E-values
 in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
-In practice, these error rates are low, and recall better than 90\% has been
+Fortunately, these error rates are low, and recall better than 90\% has been
 demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}.
@@
 The extension of this compressive acceleration approach to 
-metagenomics~\cite{yu2015entropy}, NGS read 
-mapping and chemogenomics~\cite{yorukoglu2015compressive} suggests its 
+metagenomics, NGS read 
+mapping~\cite{yorukoglu2015compressive}, and chemogenomics suggests its 
 flexibility.
+Likewise, compressive storage for these applications can be shown to scale with 
+the information-theoretic entropy of the data set~\cite{yu2015entropy}.
 

### 2015-09-22 main.tex
Put quartz first

@@
 
-Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
-Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
+Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future \cite{bonfield2013compression}.
+As stated earlier, NGS reads consist of a sequence string and associated quality scores, the latter of which generally uses more space when compressed.
+By taking advantage of biological structure, both parts of NGS reads can be better compressed.
+Unlike some other approaches to compressing quality scores in the literature
+~\cite{bonfield2013compression,ochoa2013qualcomp},
+Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
+bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
+approaches, but slightly improves the accuracy of downstream 
+variant-calling~\cite{yu2015quality}.
+Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing approaches, but slightly improves the accuracy of downstream variant-calling \cite{yu2015quality}.
+Similarly, for the sequence string,
+taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
 similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
@@
 The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
-As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
-can also be used for storage compression of NGS quality scores associated with base calls.
-Unlike many other approaches to compressing quality scores in the literature
-~\cite{bonfield2013compression,ochoa2013qualcomp},
-Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
-bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
-Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
-approaches, but slightly improves the accuracy of downstream 
-variant-calling~\cite{yu2015quality}.
 

### 2015-09-29 response.tex
Fix bad bibtex

@@
 \begin{letter}{}
-\opening{Dear Dr. Shamir,}
+\opening{Dear Ron,}
 
@@
 
-We believe we have improved this situation, not by moving references to the supplement, but by pushing the reference to the end when needed, and by citing
+We believe we have improved this situation, not by moving references to the supplement, but by pushing this reference to the end when needed, and by citing
 the original paper when possible (we freed up some space for references by narrowing the focus of the paper).
@@
 
-\emph{d. page 6 left ``compressive genomics'' is used here and is only defined later. B_D is not defined. The bottom paragraph was vague for me. }
+\emph{d. page 6 left ``compressive genomics'' is used here and is only defined later. $B_D$ is not defined. The bottom paragraph was vague for me. }
 
-We have defined B_D, rewritten the paragraph, and more properly introduced the term ``compressive genomics.''
+We have defined $B_D$, rewritten the paragraph, and more properly introduced the term ``compressive genomics.''
 
@@
 \emph{[7] The sample size argument of section 5, paragraph 2 [there are a possible
-20^500 proteins but only 6.9 x 10^7 are observed] is not persuasive. Such
+$20^500$ proteins but only $6.9 x 10^7$ are observed] is not persuasive. Such
 ``search-space-size'' arguments are nearly always vapid.  This sentence is one of

### 2015-09-29 main.tex
Fix bad bibtex

@@
 molecule end-to-end as one could read through a bound book.
-Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
+Instead, in \emph{shotgun sequencing}, DNA molecules are chopped into many small fragments;
 from these fragments we generate 
@@
 sequences about which more is known. 
-This so-called \emph{homology search} entails searching for \emph{approximate}
+This so-called \emph{homology search} entails searching for approximate
 matches in databases of known gene or protein sequences.
 The homology search problem was previously believed to be solved;
-Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} has been
+Basic Local Alignment Search Tool (\emph{BLAST})~\cite{berger2013computational} has been
 the standard tool for performing homology (similarity) search on databases of nucleotide and
@@
 that might lead to longer matches, and greedily extends them, ultimately 
-producing a \emph{sequence alignment} between a query and each potential 
+producing a sequence alignment between a query and each potential 
 database hit.
@@
 
-On a potentially even larger scale is the growth of metagenomic data.
+On a potentially even larger scale is the growth of \emph{metagenomic} data.
 Metagenomics is the study of the many genomes (bacterial, fungal, and even 
@@
 The computational study of drugs and their targets based on chemical structure and function
-is known as chemogenomics~\cite{bredel2004chemogenomics}.
+is known as \emph{chemogenomics}~\cite{bredel2004chemogenomics}.
 In the fields of drug discovery and drug repurposing, the prediction 
@@
 bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
-approaches, but slightly improves the accuracy of downstream 
-variant-calling~\cite{yu2015quality}.
 Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing approaches, but slightly improves the accuracy of downstream variant-calling \cite{yu2015quality}.
-Similarly, for the sequence string,
-taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
-similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
+Similarly, for the sequence component of the read,
+Mince~\cite{patro2015data} takes advantage of sequence redundancy by grouping
+similar reads (those that share a common short---15bp---substring) together into `buckets,' allowing that common substring to be 
 removed and treated as the bucket label, so that each read in the compressed 
